---
title: "Staurois parvus tadpole RNA-seq: differential expression analysis"
subtitle: "Androgen receptor blockade and hind-limb emergence"
author: "Kate Floer"
date: today
format:
  html:
    code-fold: true
    code-tools: true
    fig-width: 8
    fig-height: 6
    df-print: paged
execute:
  eval: false
  warning: false
  message: false
editor: source
---

::: {.callout-note collapse="true"}
## Required packages

Run once if anything below fails to load. `apeglm`, `clusterProfiler` and `org.Hs.eg.db`
are optional, the script detects their absence and either falls back to `ashr` or skips
enrichment, rather than erroring.

```r
install.packages(c("ggplot2","ggrepel","pheatmap","dplyr","tidyr","tibble","forcats",
                   "readr","stringr","RColorBrewer","matrixStats","knitr","rmarkdown"))
if (!requireNamespace("BiocManager", quietly = TRUE)) install.packages("BiocManager")
BiocManager::install(c("DESeq2","tximport","apeglm","ashr",
                       "clusterProfiler","org.Hs.eg.db","GO.db",
                       "Biostrings","Rsamtools","GenomicRanges","IRanges","pwalign"))
```
:::

```{r}
#| label: setup
#| include: false

knitr::opts_chunk$set(echo = TRUE, dev = "png", dpi = 150)

suppressPackageStartupMessages({
  library(tximport)
  library(DESeq2)
  library(ggplot2)
  library(ggrepel)
  library(pheatmap)
  library(RColorBrewer)
  library(dplyr)
  library(tidyr)
  library(tibble)
  library(readr)
  library(stringr)
  library(matrixStats)
  library(scales)      # log/pseudo-log axis transforms in §9b
})

## patchwork is only needed for the two-panel diagnostic figure in §9b.
HAVE_PATCHWORK <- requireNamespace("patchwork", quietly = TRUE)

## §9c translates the AR CDS; §9d needs forcats for factor ordering. Both degrade
## gracefully rather than stopping the render.
HAVE_BIOSTRINGS <- all(vapply(c("Biostrings", "Rsamtools", "GenomicRanges", "IRanges"),
                              requireNamespace, logical(1), quietly = TRUE))

## pairwiseAlignment(), pid() and the alignedPattern()/alignedSubject() accessors moved
## from Biostrings to pwalign and are deprecated in Biostrings >= 2.75.1. Bind them once
## from whichever package provides them, so §9c is quiet on new and old installs alike.
PW <- if (requireNamespace("pwalign", quietly = TRUE)) "pwalign" else "Biostrings"
pw_align   <- function(...) getExportedValue(PW, "pairwiseAlignment")(...)
pw_pid     <- function(...) getExportedValue(PW, "pid")(...)
pw_pattern <- function(...) getExportedValue(PW, "alignedPattern")(...)
pw_subject <- function(...) getExportedValue(PW, "alignedSubject")(...)
HAVE_FORCATS    <- requireNamespace("forcats", quietly = TRUE)

## §9d: normalise a free-text protein name to a join key against org.Hs.eg.db.
## Only two operations, both deliberately conservative:
##   1. strip a trailing "-like" / " like", which is EGAPx's marker for "resembles
##      this gene" rather than "is this gene" - the base name is the thing to look up;
##   2. lowercase and squash whitespace, because the two sources differ in case and
##      spacing but not in wording.
## Punctuation normalisation and stopword removal were both tested and REJECTED: they
## resolved 0 and 7 additional genes in the validation set respectively, so their
## precision on newly-recovered genes cannot be measured. An unauditable gain is not
## worth taking. Do not extend this function without re-running the validation in §9d.
norm_gene_name <- function(x) {
  x |> stringr::str_remove("[- ]like$") |> stringr::str_squish() |> tolower()
}

set.seed(1234)
theme_set(theme_bw(base_size = 12))
```

## Analysis parameters

All tunable settings live here so nothing is hard-coded further down.

```{r}
#| label: params

## ---- paths -----------------------------------------------------------
## PROJECT_DIR is the folder holding the salmon output and the EGAPx annotation.
##
## This does NOT hard-code "." because the working directory is not the same in
## every way of running this file: rendering sets it to the document's folder,
## but running chunks interactively leaves it wherever the R session started
## (usually the .Rproj folder, sometimes ~). A "." that works when you render
## then fails when you step through chunks, which is confusing.
##
## So: try the plausible locations and keep the first that actually contains the
## data. Add a path to the front of `candidates` if you move the folder.
PROJECT_DIR <- local({
  candidates <- c(
    ".",                                                    # render, or wd already correct
    tryCatch(dirname(rstudioapi::getSourceEditorContext()$path),  # the open document
             error = function(e) NA_character_),
    tryCatch(rstudioapi::getActiveProject(),                # the RStudio project folder
             error = function(e) NA_character_),
    "/Users/katefloer/Desktop/FrogManuscript/Ranalysis"     # known location
  )
  ## the sentinel is a small file every run needs, so a hit means a real data folder
  hit <- Filter(function(d) !is.na(d) && nzchar(d) &&
                  file.exists(file.path(d, "salmon.merged.tx2gene.tsv")),
                candidates)
  if (!length(hit)) stop(
    "Cannot locate the Ranalysis folder (looked for salmon.merged.tx2gene.tsv in: ",
    paste(Filter(function(d) !is.na(d), candidates), collapse = ", "), ").\n",
    "  Fix: set PROJECT_DIR to that folder's path in the `params` chunk.",
    call. = FALSE)
  normalizePath(hit[1])
})
message("PROJECT_DIR resolved to: ", PROJECT_DIR)
QUANT_DIR   <- file.path(PROJECT_DIR, "salmon_quant")
TX2GENE     <- file.path(PROJECT_DIR, "salmon.merged.tx2gene.tsv")
GTF_FILE    <- file.path(PROJECT_DIR, "complete.genomic.gtf")   # §9b, §10
GENOME_FA   <- file.path(PROJECT_DIR, "complete.genomic.fna")   # §9b sequence check
GENE_COUNTS <- file.path(PROJECT_DIR, "salmon.merged.gene_counts.tsv")
OUT_DIR     <- file.path(PROJECT_DIR, "results_v2")

## ---- fail loudly, and early, on a missing input ----------------------
## Without this, a wrong PROJECT_DIR surfaces 40 chunks later as an
## uninformative error from whichever function first touched the file.
local({
  required <- c(quant = QUANT_DIR, tx2gene = TX2GENE)
  missing  <- required[!file.exists(required)]
  if (length(missing))
    stop("Cannot find required input(s):\n",
         paste0("  ", names(missing), ": ", normalizePath(missing, mustWork = FALSE),
                collapse = "\n"),
         "\nSet PROJECT_DIR in the `params` chunk to your Ranalysis folder.",
         call. = FALSE)
})

## The annotation files are large and only some sections need them, so they are
## optional: the dependent sections announce themselves as skipped instead of
## breaking the render.
HAVE_GTF    <- file.exists(GTF_FILE)
HAVE_GENOME <- file.exists(GENOME_FA)

## ---- the gene the hypothesis is about --------------------------------
## Hard-coded because §9b interrogates this one model directly (exon structure,
## genomic neighbourhood, per-library reads). Confirmed from complete.genomic.gtf:
##   gene_id "egapxtmp_022918"; gene "AR"; description "androgen receptor"
AR_GENE_ID   <- "egapxtmp_022918"
AR_TX_ID     <- "egapxtmp_022918-R1"
AR_CONTIG    <- "ptg000154l"

## ---- statistical thresholds -----------------------------------------
ALPHA        <- 0.05     # FDR cutoff for calling a gene significant
LFC_CUTOFF   <- 1        # |log2FC| >= 1 (2-fold) for volcano highlighting / enrichment
MIN_COUNT    <- 10       # pre-filter: a gene needs this many counts ...
NTOP_PCA     <- 500      # top variable genes for PCA (DESeq2 plotPCA default)
NTOP_HEATMAP <- 50       # top variable genes for the gene-level heatmap

## ---- output scaffolding ---------------------------------------------
for (d in c("tables", "figures", "objects", "enrichment")) {
  dir.create(file.path(OUT_DIR, d), recursive = TRUE, showWarnings = FALSE)
}

## Consistent colours used across every figure in the report
PAL_GROUP <- c(
  young_control_body    = "#4C72B0",
  young_control_head    = "#A7C0E3",
  old_control_body      = "#55A868",
  old_control_head      = "#A8D8B4",
  old_flutamide_body    = "#C44E52",
  old_flutamide_head    = "#E8A5A7"
)
PAL_TREATMENT <- c(control = "#4C72B0", flutamide = "#C44E52")
PAL_TISSUE    <- c(body = "#8172B2", head = "#CCB974")
PAL_TIMEPOINT <- c(young = "#64B5CD", old = "#937860")
```

## 1. Sample metadata

Sample identity is parsed directly from the `salmon_quant/` directory names rather
than typed by hand, so the metadata table cannot silently drift out of sync with the
quantification files.

The naming convention is `{C|F}{45|103}-{animal}{B|H}_S{n}_L{lane}`:

- `C` / `F` &rarr; control / flutamide
- `45` / `103` &rarr; day 45 (pre-limb) / day 103 (limbs emerged)
- `B` / `H` &rarr; body (spinal cord, hind limbs, tail muscle) / head (brain, eyes, mouthparts)

```{r}
#| label: metadata

sample_dirs <- list.dirs(QUANT_DIR, full.names = FALSE, recursive = FALSE)
sample_dirs <- sample_dirs[file.exists(file.path(QUANT_DIR, sample_dirs, "quant.sf"))]
stopifnot(length(sample_dirs) > 0)

pat <- "^(C|F)(45|103)-(\\d+)(B|H)_S(\\d+)_L(\\d+)$"
stopifnot(all(grepl(pat, sample_dirs)))   # fail loudly if a name is unexpected

coldata <- tibble(sample_id = sample_dirs) |>
  mutate(
    treatment = recode(str_match(sample_id, pat)[, 2], C = "control", F = "flutamide"),
    timepoint = recode(str_match(sample_id, pat)[, 3], `45` = "young", `103` = "old"),
    animal_no = str_match(sample_id, pat)[, 4],
    tissue    = recode(str_match(sample_id, pat)[, 5], B = "body",    H = "head"),
    ## animal = the individual tadpole; head and body of one tadpole share this ID
    animal    = paste0(substr(treatment, 1, 1), "_", timepoint, "_", animal_no),
    ## group = the single combined factor used as the DESeq2 design (see below)
    group     = paste(timepoint, treatment, tissue, sep = "_")
  ) |>
  mutate(across(c(treatment, timepoint, tissue, animal, group), factor)) |>
  ## explicit reference levels: DESeq2 otherwise picks them alphabetically
  mutate(
    treatment = relevel(treatment, ref = "control"),
    timepoint = relevel(timepoint, ref = "young"),
    tissue    = relevel(tissue,    ref = "body")
  ) |>
  as.data.frame()

rownames(coldata) <- coldata$sample_id

knitr::kable(coldata[, c("sample_id", "timepoint", "treatment", "tissue", "animal", "group")],
             row.names = FALSE, caption = "Sequenced libraries and their experimental assignment")
```

### Design balance

```{r}
#| label: design-balance
group_n <- as.data.frame(table(coldata$group), responseName = "n")
names(group_n)[1] <- "group"
knitr::kable(group_n[order(-group_n$n), ], row.names = FALSE,
             caption = "Replicates per experimental group")
```

::: {.callout-important}
## Power is not uniform across groups

`old_control_head` has only **n = 2**. Both contrasts that use it, the nervous-system
developmental contrast and the head AR-blockade contrast, are substantially less powerful
than the body contrasts, and dispersion for those genes is estimated largely from the rest of
the experiment. Treat head results as exploratory, and expect fewer significant genes there
for reasons of sample size rather than biology.
:::

## 2. Import quantification with tximport

Per the DESeq2 vignette, transcript-level Salmon estimates are summarised to gene level
with `tximport`, which also produces the average-transcript-length offset that corrects
for isoform-usage shifts between samples. `DESeqDataSetFromTximport()` consumes estimated
counts, never normalised values.

```{r}
#| label: tximport

files <- file.path(QUANT_DIR, coldata$sample_id, "quant.sf")
names(files) <- coldata$sample_id
stopifnot(all(file.exists(files)))

## tx2gene: column 1 = transcript_id, column 2 = gene_id. Column 3 holds the symbol,
## which we keep separately as a gene-level annotation map.
tx2gene_full <- read_tsv(TX2GENE, show_col_types = FALSE,
                         col_names = c("transcript_id", "gene_id", "gene_name"),
                         skip = 1)

tx2gene <- tx2gene_full[, c("transcript_id", "gene_id")]

gene_annot <- tx2gene_full |>
  distinct(gene_id, gene_name) |>
  group_by(gene_id) |>
  slice(1) |>
  ungroup()

txi <- tximport(files, type = "salmon", tx2gene = tx2gene)

cat("Genes quantified:", nrow(txi$counts),
    "| Samples:", ncol(txi$counts), "\n")
```

```{r}
#| label: dds-build

dds <- DESeqDataSetFromTximport(txi, colData = coldata, design = ~ group)
```

### Why a single combined `group` factor?

The vignette recommends, for designs where the goal is "the effect of a condition within
specific subsets of samples", collapsing the factors of interest into one factor and using
`~ group`. This is equivalent to a full timepoint x treatment x tissue interaction model but
makes every comparison we want a simple, unambiguous `contrast=` call, and it avoids the
rank-deficiency problems of a fully crossed model in which flutamide exists only at day 103.

```{r}
#| label: design-note
cat("Design:", deparse(design(dds)), "\n")
cat("Levels:", paste(levels(dds$group), collapse = ", "), "\n")
```

### Pre-filtering

Following the vignette: keep genes with at least `MIN_COUNT` reads in at least as many
samples as the *smallest group*. This is a memory/speed measure, `results()` performs
independent filtering on top of it.

```{r}
#| label: prefilter

smallest_group <- min(table(coldata$group))
keep <- rowSums(counts(dds) >= MIN_COUNT) >= smallest_group

filter_summary <- data.frame(
  smallest_group_size = smallest_group,
  genes_before        = nrow(dds),
  genes_kept          = sum(keep),
  genes_removed       = sum(!keep)
)
knitr::kable(filter_summary, caption = "Pre-filtering summary")

dds <- dds[keep, ]
```

## 3. Fit the model

```{r}
#| label: deseq-fit

dds <- DESeq(dds)
resultsNames(dds)
```

```{r}
#| label: dispersion-plot
#| fig-cap: "Dispersion estimates. Black = gene-wise, red = fitted trend, blue = final shrunken values used for testing. A well-behaved experiment shows black points scattered around a smoothly decreasing red trend."

png(file.path(OUT_DIR, "figures", "dispersion_estimates.png"),
    width = 1600, height = 1300, res = 200)
plotDispEsts(dds, main = "Dispersion estimates")
invisible(dev.off())

plotDispEsts(dds, main = "Dispersion estimates")
```

### Variance-stabilising transformation

VST output is used for **all** clustering and ordination below. It is never used for
differential testing, which always operates on raw counts. `blind = FALSE` means the
dispersion trend already fitted under our design is reused; the design is not used to
remove variation, only to estimate the mean-variance trend.

```{r}
#| label: vst
vsd <- vst(dds, blind = FALSE)
saveRDS(vsd, file.path(OUT_DIR, "objects", "vsd.rds"))
saveRDS(dds, file.path(OUT_DIR, "objects", "dds_group_fitted.rds"))
```

## 4. PCA of all samples (top 500 variable genes)

```{r}
#| label: pca
#| fig-cap: "PCA on the 500 most variable genes after VST. Shape distinguishes tissue; colour distinguishes experimental group."

pca_data <- plotPCA(vsd, intgroup = c("group", "tissue", "treatment", "timepoint"),
                    ntop = NTOP_PCA, returnData = TRUE)
percent_var <- round(100 * attr(pca_data, "percentVar"))

## plotPCA() overwrites `group` with a colon-joined combination of every intgroup
## variable ("young_control_body:body:control:young"), which then matches nothing in
## PAL_GROUP and silently renders every point grey. Restore the real factors from colData.
pca_data <- pca_data |>
  select(-group) |>
  left_join(coldata |> select(name = sample_id, group), by = "name") |>
  mutate(group = factor(group, levels = names(PAL_GROUP)))

stopifnot(!any(is.na(pca_data$group)))   # fail loudly rather than plotting grey

p_pca <- ggplot(pca_data, aes(PC1, PC2, colour = group, shape = tissue)) +
  geom_point(size = 4, alpha = 0.9) +
  geom_text_repel(aes(label = name), size = 2.5, max.overlaps = 20,
                  show.legend = FALSE, colour = "grey30") +
  scale_colour_manual(values = PAL_GROUP) +
  labs(
    x = paste0("PC1: ", percent_var[1], "% variance"),
    y = paste0("PC2: ", percent_var[2], "% variance"),
    title = paste0("PCA, top ", NTOP_PCA, " most variable genes (VST)"),
    colour = "Group", shape = "Tissue"
  ) +
  coord_fixed()

ggsave(file.path(OUT_DIR, "figures", "PCA_top500.png"), plot = p_pca,
       width = 9, height = 7, dpi = 300)
write_csv(pca_data, file.path(OUT_DIR, "tables", "PCA_coordinates.csv"))
p_pca
```

```{r}
#| label: pca-facets
#| fig-cap: "The same ordination coloured by each experimental factor separately, to see which factor PC1 and PC2 actually track."

pca_long <- pca_data |>
  select(name, PC1, PC2, tissue, treatment, timepoint) |>
  pivot_longer(c(tissue, treatment, timepoint),
               names_to = "factor", values_to = "level")

p_pca_facets <- ggplot(pca_long, aes(PC1, PC2, colour = level)) +
  geom_point(size = 3) +
  facet_wrap(~ factor, nrow = 1) +
  scale_colour_manual(values = c(PAL_TISSUE, PAL_TREATMENT, PAL_TIMEPOINT)) +
  labs(x = paste0("PC1: ", percent_var[1], "%"),
       y = paste0("PC2: ", percent_var[2], "%"),
       colour = NULL, title = "PC1/PC2 by experimental factor")

ggsave(file.path(OUT_DIR, "figures", "PCA_by_factor.png"), plot = p_pca_facets,
       width = 12, height = 4.5, dpi = 300)
p_pca_facets
```

```{r}
#| label: scree
#| fig-cap: "Scree plot: variance explained by the first principal components."

vsd_mat  <- assay(vsd)
rv       <- rowVars(vsd_mat)
sel      <- order(rv, decreasing = TRUE)[seq_len(min(NTOP_PCA, length(rv)))]
pca_full <- prcomp(t(vsd_mat[sel, ]))
var_exp  <- (pca_full$sdev^2 / sum(pca_full$sdev^2)) * 100

scree_df <- data.frame(PC = factor(paste0("PC", 1:10), levels = paste0("PC", 1:10)),
                       variance = var_exp[1:10])

p_scree <- ggplot(scree_df, aes(PC, variance)) +
  geom_col(fill = "#4C72B0") +
  geom_text(aes(label = sprintf("%.1f%%", variance)), vjust = -0.4, size = 3) +
  labs(y = "Variance explained (%)", x = NULL, title = "Scree plot")

ggsave(file.path(OUT_DIR, "figures", "PCA_scree.png"), plot = p_scree,
       width = 7, height = 4.5, dpi = 300)
p_scree
```

## 5. Sample clustering

::: {.callout-note}
## All genes, or only some?: the practical answer

It depends on what is being clustered, and the two cases pull in opposite directions.

**Clustering samples** (the dendrogram you want for QC): use **all filtered genes**. The
DESeq2 vignette computes sample distances as `dist(t(assay(vsd)))` over the whole matrix.
Selecting variable genes first is a form of feature selection on the same data you then
cluster, which makes groups look cleaner than they are and can manufacture apparent
structure. Using every gene gives an honest picture of global similarity, and with ~20k
genes the distances are stable.

**Clustering genes** (the expression heatmap): use a **restricted set**, either the top
variable genes or the significant genes from a contrast. Clustering all 20k genes produces
a figure with no readable structure and is dominated by genes that do not change.

Below we do both, and additionally show the top-variable-gene sample dendrogram so you can
confirm the two agree. If they disagree, the all-gene version is the one to trust.
:::

```{r}
#| label: sample-dendrogram
#| fig-cap: "Hierarchical clustering of samples (Ward linkage on Euclidean VST distances, all filtered genes)."
#| fig-height: 7

sample_dists <- dist(t(assay(vsd)))
hc_all <- hclust(sample_dists, method = "ward.D2")

lab_cols <- PAL_GROUP[as.character(coldata[hc_all$labels, "group"])]

plot_dendro <- function(hc, cols, title) {
  op <- par(mar = c(11, 4.5, 3, 12), xpd = NA)
  on.exit(par(op))
  plot(hc, main = title, xlab = "", sub = "",
       ylab = "Euclidean distance (VST)", cex = 0.8, hang = -1)
  ## colour-code the tips by group
  ord <- hc$order
  axis(1, at = seq_along(ord), labels = FALSE, tick = FALSE)
  points(seq_along(ord), rep(par("usr")[3], length(ord)),
         col = cols[ord], pch = 15, cex = 1.4, xpd = NA)
  legend("topright", inset = c(-0.23, 0), legend = names(PAL_GROUP),
         fill = PAL_GROUP, border = NA, bty = "n", cex = 0.75, title = "Group")
}

png(file.path(OUT_DIR, "figures", "dendrogram_all_genes.png"),
    width = 2200, height = 1400, res = 190)
plot_dendro(hc_all, lab_cols, "Sample clustering: all filtered genes")
invisible(dev.off())

plot_dendro(hc_all, lab_cols, "Sample clustering: all filtered genes")
```

```{r}
#| label: sample-dendrogram-top
#| fig-cap: "The same clustering restricted to the top 500 variable genes, shown for comparison only."
#| fig-height: 7

hc_top <- hclust(dist(t(vsd_mat[sel, ])), method = "ward.D2")
lab_cols_top <- PAL_GROUP[as.character(coldata[hc_top$labels, "group"])]

png(file.path(OUT_DIR, "figures", "dendrogram_top500_genes.png"),
    width = 2200, height = 1400, res = 190)
plot_dendro(hc_top, lab_cols_top, paste0("Sample clustering: top ", NTOP_PCA, " variable genes"))
invisible(dev.off())

plot_dendro(hc_top, lab_cols_top, paste0("Sample clustering: top ", NTOP_PCA, " variable genes"))
```

```{r}
#| label: sample-distance-heatmap
#| fig-cap: "Sample-to-sample Euclidean distances on VST data (all filtered genes). Dark = similar."
#| fig-height: 8

dist_mat <- as.matrix(sample_dists)
ann_col <- coldata[, c("timepoint", "treatment", "tissue")]

p_dist <- pheatmap(
  dist_mat,
  clustering_distance_rows = sample_dists,
  clustering_distance_cols = sample_dists,
  clustering_method = "ward.D2",
  annotation_row = ann_col,
  annotation_colors = list(treatment = PAL_TREATMENT,
                           tissue = PAL_TISSUE,
                           timepoint = PAL_TIMEPOINT),
  col = colorRampPalette(rev(brewer.pal(9, "Blues")))(255),
  main = "Sample-to-sample distances (all genes)",
  fontsize = 8, silent = TRUE
)

ggsave(file.path(OUT_DIR, "figures", "sample_distance_heatmap.png"),
       plot = p_dist$gtable, width = 10, height = 8.5, dpi = 300)
grid::grid.newpage(); grid::grid.draw(p_dist$gtable)
```

```{r}
#| label: gene-heatmap
#| fig-cap: "Top 50 most variable genes (VST, centred per gene). Gene-level clustering deliberately uses a restricted gene set."
#| fig-height: 9

top_var  <- order(rowVars(vsd_mat), decreasing = TRUE)[1:NTOP_HEATMAP]
mat      <- vsd_mat[top_var, ]
mat      <- mat - rowMeans(mat)

## label rows with symbols where the annotation has one
row_lab <- gene_annot$gene_name[match(rownames(mat), gene_annot$gene_id)]
row_lab[is.na(row_lab) | row_lab == ""] <- rownames(mat)[is.na(row_lab) | row_lab == ""]
rownames(mat) <- make.unique(row_lab)

p_genes <- pheatmap(
  mat,
  annotation_col = ann_col,
  annotation_colors = list(treatment = PAL_TREATMENT,
                           tissue = PAL_TISSUE,
                           timepoint = PAL_TIMEPOINT),
  clustering_method = "ward.D2",
  scale = "none", show_colnames = TRUE,
  main = paste0("Top ", NTOP_HEATMAP, " variable genes (VST, mean-centred)"),
  fontsize = 8, fontsize_row = 7, silent = TRUE
)

ggsave(file.path(OUT_DIR, "figures", "heatmap_top_variable_genes.png"),
       plot = p_genes$gtable, width = 11, height = 10, dpi = 300)
grid::grid.newpage(); grid::grid.draw(p_genes$gtable)
```

## 6. Differential expression

### Helper functions

One function runs a contrast end-to-end: test, shrink, annotate, write CSV. A second
draws the volcano. Defining them once keeps every contrast strictly comparable.

```{r}
#| label: de-helpers

#' Run one DESeq2 contrast on the `group` factor.
#'
#' Shrinkage uses type = "ashr". This matters: the vignette's shrinkage table shows
#' apeglm does NOT support the `contrast` argument (only `coef`), while ashr does.
#' Because every comparison here is a contrast between two levels of `group`, ashr is
#' the correct estimator. Shrunken LFCs are used for ranking and plotting; p-values and
#' padj are unchanged by shrinkage.
run_contrast <- function(dds, numerator, denominator, label, description) {

  res_raw <- results(dds,
                     contrast = c("group", numerator, denominator),
                     alpha = ALPHA)

  res_shrunk <- lfcShrink(dds,
                          contrast = c("group", numerator, denominator),
                          res = res_raw, type = "ashr", quiet = TRUE)

  out <- as.data.frame(res_shrunk) |>
    rownames_to_column("gene_id") |>
    left_join(gene_annot, by = "gene_id") |>
    mutate(
      gene_name  = ifelse(is.na(gene_name) | gene_name == "", gene_id, gene_name),
      lfc_raw    = as.data.frame(res_raw)$log2FoldChange,
      contrast   = label,
      numerator  = numerator,
      denominator = denominator,
      significant = !is.na(padj) & padj < ALPHA,
      direction   = case_when(
        significant &  log2FoldChange >=  LFC_CUTOFF ~ "up",
        significant &  log2FoldChange <= -LFC_CUTOFF ~ "down",
        significant                                  ~ "significant, |LFC| < cutoff",
        TRUE                                         ~ "n.s."
      )
    ) |>
    arrange(padj, desc(abs(log2FoldChange))) |>
    select(gene_id, gene_name, baseMean, log2FoldChange, lfc_raw, lfcSE,
           pvalue, padj, significant, direction, contrast, numerator, denominator)

  write_csv(out, file.path(OUT_DIR, "tables", paste0(label, "_DE_results.csv")))

  attr(out, "description")  <- description
  attr(out, "n_tested")     <- sum(!is.na(out$padj))
  out
}

#' Volcano plot from a run_contrast() table.
plot_volcano <- function(res_df, title, subtitle = NULL, n_label = 15) {

  d <- res_df |>
    filter(!is.na(padj), !is.na(log2FoldChange)) |>
    mutate(
      neglog10p = -log10(padj),
      status = case_when(
        padj < ALPHA & log2FoldChange >=  LFC_CUTOFF ~ "Up",
        padj < ALPHA & log2FoldChange <= -LFC_CUTOFF ~ "Down",
        padj < ALPHA                                 ~ "FDR only",
        TRUE                                         ~ "n.s."
      )
    )

  ## cap infinite -log10(padj) so a single gene cannot flatten the plot
  finite_max <- suppressWarnings(max(d$neglog10p[is.finite(d$neglog10p)], na.rm = TRUE))
  if (!is.finite(finite_max)) finite_max <- 1
  d$neglog10p <- pmin(d$neglog10p, finite_max * 1.05)

  ## Genes detected in essentially one group produce |log2FC| of 15-25 even after
  ## shrinkage. They are real signals but rest on very few reads, so they are labelled
  ## with an asterisk rather than presented on the same footing as well-expressed genes.
  d <- d |> mutate(low_expr = baseMean < 50)

  lab <- d |>
    filter(status %in% c("Up", "Down")) |>
    arrange(padj, desc(abs(log2FoldChange))) |>
    head(n_label) |>
    mutate(gene_label = ifelse(low_expr, paste0(gene_name, " *"), gene_name))

  n_up   <- sum(d$status == "Up")
  n_down <- sum(d$status == "Down")

  ggplot(d, aes(log2FoldChange, neglog10p)) +
    geom_point(aes(colour = status), size = 1.3, alpha = 0.65) +
    geom_vline(xintercept = c(-LFC_CUTOFF, LFC_CUTOFF),
               linetype = "dashed", colour = "grey45", linewidth = 0.35) +
    geom_hline(yintercept = -log10(ALPHA),
               linetype = "dashed", colour = "grey45", linewidth = 0.35) +
    geom_text_repel(data = lab, aes(label = gene_label),
                    size = 3, max.overlaps = 25, min.segment.length = 0,
                    segment.colour = "grey55", box.padding = 0.4) +
    scale_colour_manual(values = c(Up = "#C44E52", Down = "#4C72B0",
                                   `FDR only` = "#DD8452", n.s. = "grey80")) +
    labs(
      title = title,
      subtitle = subtitle %||% sprintf(
        "%d up, %d down at FDR < %.2f and |log2FC| >= %g   (%d genes tested)",
        n_up, n_down, ALPHA, LFC_CUTOFF, nrow(d)),
      x = "log2 fold change (ashr-shrunken)",
      y = "-log10 adjusted p-value",
      colour = NULL,
      caption = if (any(lab$low_expr)) "* baseMean < 50: large fold change rests on few reads" else NULL
    ) +
    theme(legend.position = "top",
          plot.caption = element_text(size = 8, colour = "grey35", hjust = 0))
}

`%||%` <- function(a, b) if (is.null(a)) b else a

#' Compact summary row for a contrast.
summarise_contrast <- function(res_df) {
  data.frame(
    contrast   = unique(res_df$contrast),
    genes_tested = sum(!is.na(res_df$padj)),
    sig_FDR      = sum(res_df$significant, na.rm = TRUE),
    up           = sum(res_df$direction == "up"),
    down         = sum(res_df$direction == "down")
  )
}
```

### 6a. Developmental change with leg emergence: body

**old_control_body vs young_control_body** (n = 4 vs 5). Captures everything that changes in
the trunk/hind-limb/tail compartment between day 45 and day 103, including limb outgrowth
itself.

```{r}
#| label: de-body-dev
res_body_dev <- run_contrast(
  dds, "old_control_body", "young_control_body",
  label = "leg_emergence_body_old_vs_young",
  description = "Developmental change in body: day 103 vs day 45, controls only"
)
knitr::kable(head(res_body_dev, 20), digits = 3,
             caption = "Top 20 genes - leg emergence (body)")
```

```{r}
#| label: volcano-body-dev
#| fig-cap: "Volcano: old vs young control body."
p1 <- plot_volcano(res_body_dev, "Leg emergence: old vs young control BODY")
ggsave(file.path(OUT_DIR, "figures", "volcano_leg_emergence_body.png"),
       plot = p1, width = 8.5, height = 7, dpi = 300)
p1
```

### 6b. Developmental change in the nervous system: head

**old_control_head vs young_control_head** (n = 2 vs 4). Underpowered on the old side; read
with the caveat above.

```{r}
#| label: de-head-dev
res_head_dev <- run_contrast(
  dds, "old_control_head", "young_control_head",
  label = "nervous_system_head_old_vs_young",
  description = "Developmental change in head: day 103 vs day 45, controls only"
)
knitr::kable(head(res_head_dev, 20), digits = 3,
             caption = "Top 20 genes - nervous system development (head)")
```

```{r}
#| label: volcano-head-dev
#| fig-cap: "Volcano: old vs young control head."
p2 <- plot_volcano(res_head_dev, "Nervous system development - old vs young control HEAD")
ggsave(file.path(OUT_DIR, "figures", "volcano_nervous_system_head.png"),
       plot = p2, width = 8.5, height = 7, dpi = 300)
p2
```

### 6c. AR blockade in older tadpoles: body

**old_flutamide_body vs old_control_body** (n = 5 vs 4). The primary test of androgen
receptor involvement in the limb/trunk compartment.

```{r}
#| label: de-body-flu
res_body_flu <- run_contrast(
  dds, "old_flutamide_body", "old_control_body",
  label = "AR_blockade_body_flutamide_vs_control",
  description = "Flutamide vs control, day 103, body"
)
knitr::kable(head(res_body_flu, 20), digits = 3,
             caption = "Top 20 genes - AR blockade (body)")
```

```{r}
#| label: volcano-body-flu
#| fig-cap: "Volcano: flutamide vs control, old body."
p3 <- plot_volcano(res_body_flu, "AR blockade: flutamide vs control, old BODY")
ggsave(file.path(OUT_DIR, "figures", "volcano_AR_blockade_body.png"),
       plot = p3, width = 8.5, height = 7, dpi = 300)
p3
```

### 6d. AR blockade in older tadpoles: head

**old_flutamide_head vs old_control_head** (n = 4 vs 2).

```{r}
#| label: de-head-flu
res_head_flu <- run_contrast(
  dds, "old_flutamide_head", "old_control_head",
  label = "AR_blockade_head_flutamide_vs_control",
  description = "Flutamide vs control, day 103, head"
)
knitr::kable(head(res_head_flu, 20), digits = 3,
             caption = "Top 20 genes - AR blockade (head)")
```

```{r}
#| label: volcano-head-flu
#| fig-cap: "Volcano: flutamide vs control, old head."
p4 <- plot_volcano(res_head_flu, "AR blockade: flutamide vs control, old HEAD")
ggsave(file.path(OUT_DIR, "figures", "volcano_AR_blockade_head.png"),
       plot = p4, width = 8.5, height = 7, dpi = 300)
p4
```

## 7. Whole-animal AR blockade: old control vs old flutamide

You asked to collapse head and body into one sample per animal. There are two defensible
ways to do that and they answer slightly different questions, so both are run.

::: {.callout-warning}
## `collapseReplicates` is for technical replicates only

The DESeq2 vignette is explicit that `collapseReplicates` is intended for multiple
sequencing runs of the *same library*, and that biological replicates must not be collapsed.
Head and body are different tissues, not technical replicates, but summing their counts is
still a legitimate operation if what you want is a **whole-animal transcriptome**. It is a
change of biological unit, not a technical merge, so we treat it as such and interpret it
that way: fold changes become "per whole tadpole", weighted by how much RNA each compartment
contributes.

The harder problem is that only **6 of 9** day-103 animals have both tissues sequenced
(controls C103-4, C103-5; flutamide F103-1, F103-2, F103-3, F103-5). Summing an animal that
is missing its head would compare a body-only pseudo-animal against whole animals, a
composition artifact masquerading as a treatment effect. Approach A therefore uses only
complete animals, which costs power (**2 vs 4**). Approach B keeps every sample and models
tissue as a covariate instead, which is the statistically stronger option.
:::

### Approach A: sum head + body per animal (complete animals only)

```{r}
#| label: collapse-complete

old_meta <- coldata[coldata$timepoint == "old", ]
tissue_per_animal <- table(old_meta$animal, old_meta$tissue)
complete_animals  <- rownames(tissue_per_animal)[rowSums(tissue_per_animal > 0) == 2]

cat("Day-103 animals with both tissues:", paste(complete_animals, collapse = ", "), "\n")
cat("Excluded (single tissue only):",
    paste(setdiff(unique(as.character(old_meta$animal)), complete_animals), collapse = ", "), "\n")

keep_cols <- rownames(coldata)[coldata$animal %in% complete_animals]

dds_pair <- dds[, keep_cols]
dds_pair$animal <- droplevels(dds_pair$animal)

## Sum counts across the two tissues of each animal.
dds_collapsed <- collapseReplicates(dds_pair,
                                    groupby = dds_pair$animal,
                                    run = dds_pair$sample_id)

colData(dds_collapsed)$treatment <- relevel(droplevels(colData(dds_collapsed)$treatment),
                                            ref = "control")
design(dds_collapsed) <- ~ treatment

## sanity check: collapsed counts must equal the sum of their parts
chk_animal <- complete_animals[1]
chk <- all.equal(
  as.numeric(counts(dds_collapsed)[, chk_animal]),
  as.numeric(rowSums(counts(dds_pair)[, dds_pair$animal == chk_animal, drop = FALSE]))
)
stopifnot(isTRUE(chk))

knitr::kable(as.data.frame(colData(dds_collapsed)[, c("treatment", "runsCollapsed")]),
             caption = "Whole-animal pseudo-samples")

dds_collapsed <- DESeq(dds_collapsed)

res_whole <- results(dds_collapsed, contrast = c("treatment", "flutamide", "control"),
                     alpha = ALPHA)
res_whole_shrunk <- lfcShrink(dds_collapsed, contrast = c("treatment", "flutamide", "control"),
                              res = res_whole, type = "ashr", quiet = TRUE)

res_whole_df <- as.data.frame(res_whole_shrunk) |>
  rownames_to_column("gene_id") |>
  left_join(gene_annot, by = "gene_id") |>
  mutate(gene_name = ifelse(is.na(gene_name) | gene_name == "", gene_id, gene_name),
         contrast = "whole_animal_collapsed_flutamide_vs_control",
         significant = !is.na(padj) & padj < ALPHA,
         direction = case_when(
           significant & log2FoldChange >=  LFC_CUTOFF ~ "up",
           significant & log2FoldChange <= -LFC_CUTOFF ~ "down",
           significant ~ "significant, |LFC| < cutoff", TRUE ~ "n.s.")) |>
  arrange(padj)

write_csv(res_whole_df,
          file.path(OUT_DIR, "tables", "whole_animal_collapsed_flutamide_vs_control_DE_results.csv"))

knitr::kable(head(res_whole_df, 20), digits = 3,
             caption = "Top 20 genes - whole-animal (collapsed), flutamide vs control")
```

```{r}
#| label: volcano-whole
#| fig-cap: "Volcano: whole-animal collapsed comparison (2 control vs 4 flutamide)."
p5 <- plot_volcano(res_whole_df,
                   "AR blockade: whole animal (head + body summed)",
                   subtitle = "Complete animals only: 2 control vs 4 flutamide - interpret cautiously")
ggsave(file.path(OUT_DIR, "figures", "volcano_whole_animal_collapsed.png"),
       plot = p5, width = 8.5, height = 7, dpi = 300)
p5
```

### Approach B: all day-103 samples, tissue as a covariate

`~ tissue + treatment` uses all 15 day-103 libraries and estimates the flutamide effect
*after* absorbing the large baseline difference between head and body. This is the
recommended primary analysis for "does flutamide change expression in older tadpoles,
overall".

```{r}
#| label: old-tissue-covariate

dds_old <- dds[, coldata$timepoint == "old"]
colData(dds_old)$treatment <- relevel(droplevels(colData(dds_old)$treatment), ref = "control")
colData(dds_old)$tissue    <- droplevels(colData(dds_old)$tissue)
design(dds_old) <- ~ tissue + treatment

dds_old <- DESeq(dds_old)
cat("Coefficients:", paste(resultsNames(dds_old), collapse = " | "), "\n")

res_old <- results(dds_old, name = "treatment_flutamide_vs_control", alpha = ALPHA)

## apeglm is the preferred estimator here because this is a coefficient, not a contrast.
## If apeglm is not installed we fall back to ashr, which is also valid for a coef.
shrink_type <- if (requireNamespace("apeglm", quietly = TRUE)) "apeglm" else "ashr"
cat("Shrinkage estimator used:", shrink_type, "\n")

res_old_shrunk <- lfcShrink(dds_old, coef = "treatment_flutamide_vs_control",
                            type = shrink_type, res = res_old, quiet = TRUE)

res_old_df <- as.data.frame(res_old_shrunk) |>
  rownames_to_column("gene_id") |>
  left_join(gene_annot, by = "gene_id") |>
  mutate(gene_name = ifelse(is.na(gene_name) | gene_name == "", gene_id, gene_name),
         contrast = "old_all_tissues_flutamide_vs_control_tissue_adjusted",
         significant = !is.na(padj) & padj < ALPHA,
         direction = case_when(
           significant & log2FoldChange >=  LFC_CUTOFF ~ "up",
           significant & log2FoldChange <= -LFC_CUTOFF ~ "down",
           significant ~ "significant, |LFC| < cutoff", TRUE ~ "n.s.")) |>
  arrange(padj)

write_csv(res_old_df,
          file.path(OUT_DIR, "tables", "old_all_tissues_flutamide_vs_control_DE_results.csv"))

knitr::kable(head(res_old_df, 20), digits = 3,
             caption = "Top 20 genes - day 103 flutamide vs control, tissue-adjusted")
```

```{r}
#| label: volcano-old-adjusted
#| fig-cap: "Volcano: day-103 flutamide vs control, adjusted for tissue (all 15 libraries)."
p6 <- plot_volcano(res_old_df,
                   "AR blockade: all day-103 samples, tissue-adjusted",
                   subtitle = "Design ~ tissue + treatment, n = 15 libraries")
ggsave(file.path(OUT_DIR, "figures", "volcano_old_tissue_adjusted.png"),
       plot = p6, width = 8.5, height = 7, dpi = 300)
p6
```

### Does flutamide act differently in head vs body?

An interaction test asks directly whether the AR-blockade response differs by tissue.

```{r}
#| label: interaction-test

dds_int <- dds[, coldata$timepoint == "old"]
colData(dds_int)$treatment <- relevel(droplevels(colData(dds_int)$treatment), ref = "control")
colData(dds_int)$tissue    <- droplevels(colData(dds_int)$tissue)
design(dds_int) <- ~ tissue + treatment + tissue:treatment

dds_int <- DESeq(dds_int, test = "LRT", reduced = ~ tissue + treatment)
res_int <- results(dds_int, alpha = ALPHA)

res_int_df <- as.data.frame(res_int) |>
  rownames_to_column("gene_id") |>
  left_join(gene_annot, by = "gene_id") |>
  mutate(gene_name = ifelse(is.na(gene_name) | gene_name == "", gene_id, gene_name)) |>
  arrange(padj)

write_csv(res_int_df,
          file.path(OUT_DIR, "tables", "tissue_by_treatment_interaction_LRT.csv"))

cat("Genes with a significant tissue x treatment interaction (LRT, FDR <",
    ALPHA, "):", sum(res_int_df$padj < ALPHA, na.rm = TRUE), "\n")

knitr::kable(head(res_int_df, 15), digits = 3,
             caption = "Top interaction genes (LRT: does the flutamide effect differ by tissue?)")
```

## 8. Contrast summary

```{r}
#| label: contrast-summary

all_summaries <- bind_rows(
  summarise_contrast(res_body_dev),
  summarise_contrast(res_head_dev),
  summarise_contrast(res_body_flu),
  summarise_contrast(res_head_flu),
  data.frame(contrast = "whole_animal_collapsed_flutamide_vs_control",
             genes_tested = sum(!is.na(res_whole_df$padj)),
             sig_FDR = sum(res_whole_df$significant),
             up = sum(res_whole_df$direction == "up"),
             down = sum(res_whole_df$direction == "down")),
  data.frame(contrast = "old_all_tissues_flutamide_vs_control_tissue_adjusted",
             genes_tested = sum(!is.na(res_old_df$padj)),
             sig_FDR = sum(res_old_df$significant),
             up = sum(res_old_df$direction == "up"),
             down = sum(res_old_df$direction == "down"))
)

write_csv(all_summaries, file.path(OUT_DIR, "tables", "contrast_summary.csv"))
knitr::kable(all_summaries, caption = "Significant genes per contrast")
```

```{r}
#| label: summary-barplot
#| fig-cap: "Number of differentially expressed genes per contrast."

sum_long <- all_summaries |>
  select(contrast, up, down) |>
  pivot_longer(c(up, down), names_to = "direction", values_to = "n") |>
  mutate(n_signed = ifelse(direction == "down", -n, n),
         contrast = factor(contrast, levels = rev(all_summaries$contrast)))

p_sum <- ggplot(sum_long, aes(n_signed, contrast, fill = direction)) +
  geom_col() +
  geom_vline(xintercept = 0, colour = "grey30") +
  geom_text(aes(label = n, hjust = ifelse(sum_long$direction == "down", 1.2, -0.2)),
            size = 3) +
  scale_fill_manual(values = c(up = "#C44E52", down = "#4C72B0")) +
  labs(x = paste0("Genes (FDR < ", ALPHA, ", |log2FC| >= ", LFC_CUTOFF, ")"),
       y = NULL, fill = NULL, title = "Differentially expressed genes by contrast")

ggsave(file.path(OUT_DIR, "figures", "DE_gene_counts_by_contrast.png"),
       plot = p_sum, width = 10, height = 5, dpi = 300)
p_sum
```

## 9. Androgen signalling genes

A targeted look at the pathway the hypothesis is about, independent of whether these genes
survive genome-wide correction.

### Detection check: read this before interpreting anything below

Pre-filtering silently removes genes, so before asking *"is AR differentially expressed?"*
we must ask *"is AR detected at all?"*. This table reports raw counts for every pathway gene
**before** filtering.

```{r}
#| label: ar-detection

ar_symbols <- c("AR", "SRD5A1", "SRD5A2", "CYP19A1", "CYP17A1", "HSD17B1", "HSD17B3",
                "STAR", "SHBG", "NR3C1", "ESR1", "ESR2", "FOXO1", "MYOD1", "MYOG")

## unfiltered counts straight from tximport
raw_counts <- txi$counts
ar_map <- gene_annot |> filter(gene_name %in% ar_symbols, gene_id %in% rownames(raw_counts))

ar_detect <- ar_map |>
  rowwise() |>
  mutate(
    max_count      = round(max(raw_counts[gene_id, ]), 1),
    median_count   = round(median(raw_counts[gene_id, ]), 1),
    n_samples_ge10 = sum(raw_counts[gene_id, ] >= MIN_COUNT),
    passed_filter  = gene_id %in% rownames(dds)
  ) |>
  ungroup() |>
  arrange(desc(n_samples_ge10))

write_csv(ar_detect, file.path(OUT_DIR, "tables", "androgen_pathway_detection.csv"))
knitr::kable(ar_detect, caption = "Detection of androgen-pathway genes BEFORE pre-filtering")

undetected <- ar_detect$gene_name[!ar_detect$passed_filter]
```

```{r}
#| label: ar-warning
#| results: asis
#| echo: false

if ("AR" %in% undetected) {
  cat("::: {.callout-important}\n")
  cat("## The androgen receptor gene is essentially not detected\n\n")
  cat("`AR` (", ar_map$gene_id[ar_map$gene_name == "AR"],
      ") has a maximum of **",
      ar_detect$max_count[ar_detect$gene_name == "AR"],
      " reads** in any single library and a median of 0. ",
      "It does not survive pre-filtering and therefore has no fold change or p-value in any ",
      "contrast in this report.\n\n", sep = "")
  cat("This is a statement about the *data*, not yet about androgen biology. There are three ",
      "candidate explanations, and Section 9b below tests each one against the evidence ",
      "rather than leaving them open:\n\n")
  cat("1. **Annotation failure** — a fragmented or truncated EGAPx gene model, so reads that ",
      "belong to AR are assigned elsewhere or not at all.\n")
  cat("2. **Quantification/index mismatch** — the Salmon index was built from a different ",
      "annotation than `salmon.merged.tx2gene.tsv`.\n")
  cat("3. **Genuinely low expression** in whole head/body homogenate at these stages. AR ",
      "transcript is often restricted to specific motor nuclei and target muscles; bulk tissue ",
      "dilutes it below detection. Note that flutamide can still act on protein translated ",
      "from transcript too sparse to quantify here.\n\n")
  cat("Regardless of which holds, the AR-blockade contrasts should be read as *the downstream ",
      "transcriptional consequences of flutamide*, not as evidence about AR expression itself.\n")
  cat(":::\n\n")
}

if (length(setdiff(undetected, "AR")) > 0) {
  cat("Also removed by pre-filtering (too few counts): **",
      paste(setdiff(undetected, "AR"), collapse = ", "), "**.\n\n", sep = "")
}
```

```{r}
#| label: ar-genes
#| fig-cap: "Normalised counts for the androgen-signalling genes that survived pre-filtering."
#| fig-height: 9

ar_ids <- gene_annot |>
  filter(gene_name %in% ar_symbols, gene_id %in% rownames(dds))

norm_counts <- counts(dds, normalized = TRUE)

ar_long <- norm_counts[ar_ids$gene_id, , drop = FALSE] |>
  as.data.frame() |>
  rownames_to_column("gene_id") |>
  pivot_longer(-gene_id, names_to = "sample_id", values_to = "norm_count") |>
  left_join(ar_ids, by = "gene_id") |>
  left_join(coldata, by = "sample_id")

p_ar <- ggplot(ar_long, aes(group, norm_count + 1, colour = treatment, shape = tissue)) +
  geom_point(position = position_jitter(width = 0.18), size = 2, alpha = 0.85) +
  facet_wrap(~ gene_name, scales = "free_y", ncol = 4) +
  scale_y_log10() +
  scale_colour_manual(values = PAL_TREATMENT) +
  labs(y = "Normalised count + 1 (log10)", x = NULL,
       title = "Androgen signalling pathway genes") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 7),
        strip.text = element_text(face = "bold"))

ggsave(file.path(OUT_DIR, "figures", "androgen_pathway_genes.png"),
       plot = p_ar, width = 13, height = 10, dpi = 300)
p_ar
```

```{r}
#| label: ar-stats

## Match on gene_id, not gene_name: the results tables substitute the gene_id as a
## display label whenever a symbol is missing, so filtering on gene_name alone would
## silently drop genes and produce an empty table.
ar_stats <- bind_rows(
  res_body_dev, res_head_dev, res_body_flu, res_head_flu
) |>
  filter(gene_id %in% ar_map$gene_id) |>
  left_join(ar_map |> rename(symbol = gene_name), by = "gene_id") |>
  select(symbol, gene_id, contrast, baseMean, log2FoldChange, lfcSE, pvalue, padj, significant) |>
  arrange(symbol, contrast)

write_csv(ar_stats, file.path(OUT_DIR, "tables", "androgen_pathway_statistics.csv"))

if (nrow(ar_stats) == 0) {
  cat("No androgen-pathway genes survived pre-filtering — see the detection table above.\n")
} else {
  knitr::kable(ar_stats, digits = 3,
               caption = "Androgen-pathway genes across the four primary contrasts")
}
```

### Is there an unnamed gene model that behaves like AR?

If the EGAPx `AR` model is broken, the real AR reads may have landed on an unnamed
`egapxtmp_*` model. This is not a substitute for a sequence-level check, but a
co-expression scan is cheap and can nominate candidates: a true AR model should be
expressed, and should respond to flutamide in a similar way to known AR-responsive genes.

```{r}
#| label: ar-candidate-scan

## Genes with a known androgen connection that ARE detected here, used as a bait set.
bait <- ar_ids$gene_id[ar_ids$gene_name %in%
                         c("SRD5A1", "SRD5A2", "CYP17A1", "NR3C1", "FOXO1")]

if (length(bait) >= 2) {

  vmat <- assay(vsd)
  bait_profile <- colMeans(vmat[bait, , drop = FALSE])

  ## correlate every gene against the mean bait profile
  cors <- cor(t(vmat), bait_profile, method = "spearman")

  ar_candidates <- tibble(
    gene_id = rownames(vmat),
    rho     = as.numeric(cors)
  ) |>
    left_join(gene_annot, by = "gene_id") |>
    filter(grepl("^egapxtmp_", gene_name) | is.na(gene_name)) |>
    arrange(desc(rho)) |>
    head(25) |>
    left_join(
      res_body_flu |> select(gene_id, flu_body_LFC = log2FoldChange, flu_body_padj = padj),
      by = "gene_id"
    )

  write_csv(ar_candidates, file.path(OUT_DIR, "tables", "unnamed_genes_correlated_with_androgen_bait.csv"))
  knitr::kable(ar_candidates, digits = 3,
               caption = paste("Unnamed gene models most correlated with the detected",
                               "androgen-pathway bait set. Candidates only — confirm by",
                               "aligning the model's sequence to a known AR protein."))
} else {
  cat("Too few detected bait genes for a co-expression scan.\n")
}
```

## 9b. Why is AR not detected?: diagnosing the three hypotheses

Section 9 established *that* `AR` is undetected. This section establishes *why*, by testing
each of the three candidate explanations against evidence that is already on disk. The logic
is eliminative: annotation failure and index mismatch both make predictions that can be
checked directly, and if both are falsified while the quantification demonstrably works for
comparable genes, what remains is a statement about the biology and the detection limit.

Everything here reads `complete.genomic.gtf` and the raw `quant.sf` files. If the GTF is not
present the section skips itself rather than failing the render.

```{r}
#| label: ar-diag-setup

## Section 9b builds on earlier chunks; say so plainly rather than failing later
## on a bare "object not found". A full render never trips this.
if (!all(vapply(c("GTF_FILE", "HAVE_GTF", "raw_counts", "gene_annot"), exists, logical(1))))
  stop("Section 9b builds on earlier chunks. Use Run > Run All Chunks Above ",
       "(Cmd+Option+P), or render the whole document.", call. = FALSE)

## This whole section is conditional on the GTF being available.
RUN_AR_DIAG <- HAVE_GTF

if (!RUN_AR_DIAG) {
  message("complete.genomic.gtf not found - skipping Section 9b.")
}

## Small helper: pull a named GTF attribute out of the 9th column.
gtf_attr <- function(x, key) {
  m <- str_match(x, paste0(key, ' "([^"]*)"'))[, 2]
  m
}
```

### Hypothesis 1: is the gene model broken?

A fragmented or truncated model would show too few exons, a short CDS, or a protein well
below the ~770-820 aa expected for an anuran androgen receptor.

```{r}
#| label: ar-model-structure

if (RUN_AR_DIAG) {

  ## grep is far cheaper than parsing a 500 MB GTF in R
  ar_lines <- system2("grep", c(shQuote(AR_GENE_ID), shQuote(GTF_FILE)), stdout = TRUE)

  ar_gtf <- read.delim(text = ar_lines, header = FALSE, sep = "\t",
                       quote = "", stringsAsFactors = FALSE,
                       col.names = c("seqname", "source", "feature", "start", "end",
                                     "score", "strand", "frame", "attribute"))

  ar_exons <- ar_gtf |> filter(feature == "exon")
  ar_cds   <- ar_gtf |> filter(feature == "CDS")

  ar_model <- tibble(
    gene_id      = AR_GENE_ID,
    contig       = ar_gtf$seqname[1],
    start        = min(ar_gtf$start),
    end          = max(ar_gtf$end),
    strand       = ar_gtf$strand[1],
    n_exons      = nrow(ar_exons),
    tx_length_nt = sum(ar_exons$end - ar_exons$start + 1),
    cds_nt       = sum(ar_cds$end - ar_cds$start + 1),
    protein_aa   = sum(ar_cds$end - ar_cds$start + 1) / 3 - 1,
    has_start    = any(ar_gtf$feature == "start_codon"),
    has_stop     = any(ar_gtf$feature == "stop_codon")
  )

  knitr::kable(ar_model, caption = "Structure of the annotated AR gene model")
}
```

```{r}
#| label: ar-model-verdict
#| results: asis
#| echo: false

if (RUN_AR_DIAG) {
  ok <- ar_model$protein_aa > 700 && ar_model$has_start && ar_model$has_stop
  cat("**Verdict:** the model encodes a **", round(ar_model$protein_aa), " aa** protein from ",
      ar_model$n_exons, " exons with ",
      if (ok) "both a start and a stop codon annotated" else "an incomplete reading frame",
      ". Anuran AR is ~770-820 aa, so this model is **",
      if (ok) "structurally complete - Hypothesis 1 is not supported" else "suspect",
      "**.\n\n", sep = "")
}
```

### Hypothesis 2: did Salmon ever see the transcript?

An index/annotation mismatch would mean the AR transcript is absent from `quant.sf`, or
present with an implausible effective length. We also read out the per-library counts here,
which the rest of the section uses.

```{r}
#| label: ar-quant-evidence

if (RUN_AR_DIAG) {

  ## For each library: the AR row, the library size, and the TPM normalising constant.
  ## Salmon's own relation is  reads_i = TPM_i * effLen_i * N / sum_j(TPM_j * effLen_j),
  ## so the denominator lets us convert "1 TPM" into "expected reads" for THIS library.
  ar_quant <- lapply(coldata$sample_id, function(s) {
    q <- readr::read_tsv(file.path(QUANT_DIR, s, "quant.sf"),
                         show_col_types = FALSE, progress = FALSE)
    row <- q |> filter(Name == AR_TX_ID)
    tibble(
      sample_id        = s,
      in_index         = nrow(row) == 1,
      eff_length       = if (nrow(row) == 1) row$EffectiveLength else NA_real_,
      AR_reads         = if (nrow(row) == 1) row$NumReads        else NA_real_,
      AR_TPM           = if (nrow(row) == 1) row$TPM             else NA_real_,
      library_size     = sum(q$NumReads),
      tpm_denominator  = sum(q$TPM * q$EffectiveLength)
    )
  }) |>
    bind_rows() |>
    left_join(coldata |> select(sample_id, group, tissue, treatment, timepoint),
              by = "sample_id") |>
    mutate(
      ## how many reads a transcript of AR's length would earn at exactly 1 TPM
      reads_per_TPM = eff_length * library_size / tpm_denominator,
      implied_TPM   = AR_reads / reads_per_TPM
    )

  write_csv(ar_quant |> select(sample_id, group, tissue, library_size,
                               eff_length, AR_reads, AR_TPM, reads_per_TPM),
            file.path(OUT_DIR, "tables", "AR_per_sample_detection.csv"))

  knitr::kable(
    ar_quant |>
      select(sample_id, group, library_size, AR_reads, AR_TPM, reads_per_TPM) |>
      mutate(library_size = round(library_size), reads_per_TPM = round(reads_per_TPM, 1)),
    digits = 4,
    caption = "AR transcript in every library, with the reads expected at 1 TPM")
}
```

```{r}
#| label: ar-quant-verdict
#| results: asis
#| echo: false

if (RUN_AR_DIAG) {
  cat("**Verdict:** the AR transcript is present in **", sum(ar_quant$in_index), "/",
      nrow(ar_quant), "** `quant.sf` files with an effective length of ",
      round(min(ar_quant$eff_length)), "-", round(max(ar_quant$eff_length)),
      " nt. Salmon had the transcript in its index and assigned it ",
      sum(ar_quant$AR_reads), " reads in total. **Hypothesis 2 is falsified** - this is not ",
      "an index or annotation mismatch.\n\n", sep = "")

  cat("The reads also do not scale with sequencing depth: the deepest library (",
      format(round(max(ar_quant$library_size)), big.mark = ","), " reads) carries ",
      ar_quant$AR_reads[which.max(ar_quant$library_size)], " AR reads, while ",
      sum(ar_quant$AR_reads == 0), " of ", nrow(ar_quant),
      " libraries carry none. Undersampling would not produce that pattern.\n\n", sep = "")
}
```

### Is the locus itself dead?

If the contig carrying AR were poorly assembled or poorly covered, every gene on it would be
silent. Testing that separates a regional artefact from a gene-specific result.

```{r}
#| label: ar-neighbourhood

if (RUN_AR_DIAG) {

  ## all gene-level records on the AR contig
  contig_cmd <- sprintf("awk -F'\\t' '$3==\"gene\" && $1==\"%s\"' %s",
                        AR_CONTIG, shQuote(GTF_FILE))
  contig_lines <- system(contig_cmd, intern = TRUE)

  contig_genes <- tibble(raw = contig_lines) |>
    mutate(
      start   = as.integer(str_split_i(raw, "\t", 4)),
      end     = as.integer(str_split_i(raw, "\t", 5)),
      gene_id = gtf_attr(raw, "gene_id"),
      symbol  = gtf_attr(raw, "gene")
    ) |>
    select(-raw) |>
    arrange(start)

  ## attach expression (raw_counts comes from the detection chunk in Section 9)
  gene_max <- apply(raw_counts, 1, max)
  gene_med <- apply(raw_counts, 1, median)
  contig_genes <- contig_genes |>
    mutate(
      max_count    = unname(gene_max[gene_id]),
      median_count = unname(gene_med[gene_id]),
      expressed    = !is.na(max_count) & max_count >= MIN_COUNT
    )

  n_expr <- sum(contig_genes$expressed)
  pct_expr <- round(100 * n_expr / nrow(contig_genes))

  ## AR's immediate neighbourhood
  ar_idx <- which(contig_genes$gene_id == AR_GENE_ID)
  nb <- contig_genes[max(1, ar_idx - 5):min(nrow(contig_genes), ar_idx + 5), ] |>
    mutate(is_AR = gene_id == AR_GENE_ID)

  knitr::kable(
    nb |> select(start, end, gene_id, symbol, median_count, max_count, is_AR),
    caption = sprintf("AR's genomic neighbourhood on %s (%d of %d genes on this contig reach %d reads: %d%%)",
                      AR_CONTIG, n_expr, nrow(contig_genes), MIN_COUNT, pct_expr))
}
```

```{r}
#| label: ar-neighbourhood-verdict
#| results: asis
#| echo: false

if (RUN_AR_DIAG) {
  cat("**Verdict:** ", n_expr, " of ", nrow(contig_genes), " gene models on `", AR_CONTIG,
      "` (", pct_expr, "%) reach ", MIN_COUNT,
      " reads, and AR's immediate neighbours are robustly expressed. The locus is ",
      "transcriptionally active; there is **no regional dropout**.\n\n", sep = "")
}
```

### The decisive comparison: AR against the other nuclear receptors

If quantification were broken, other receptors of similar length and expression class would
fail too. If instead the whole sex-steroid axis is silent while every other nuclear receptor
quantifies normally, the pattern is biological.

```{r}
#| label: ar-receptor-panel

if (RUN_AR_DIAG) {

  receptor_panel <- tribble(
    ~symbol,     ~class,
    "GAPDH",     "reference",
    "THRA",      "other nuclear receptor",
    "THRB",      "other nuclear receptor",
    "RARA",      "other nuclear receptor",
    "RARB",      "other nuclear receptor",
    "VDR",       "other nuclear receptor",
    "ESRRA",     "other nuclear receptor",
    "NR3C1",     "other nuclear receptor",
    "NR3C2",     "other nuclear receptor",
    "CYP17A1",   "other nuclear receptor",
    "AR",        "androgen receptor",
    "PGR",       "sex-steroid axis",
    "ESR1",      "sex-steroid axis",
    "ESR2",      "sex-steroid axis",
    "CYP19A1",   "sex-steroid axis",
    "STAR",      "sex-steroid axis",
    "SHBG",      "sex-steroid axis",
    "HSD17B3",   "sex-steroid axis",
    "SRD5A1",    "sex-steroid axis",
    "SRD5A2",    "sex-steroid axis"
  )

  panel_expr <- receptor_panel |>
    left_join(gene_annot |> select(gene_id, gene_name), by = c("symbol" = "gene_name")) |>
    filter(!is.na(gene_id), gene_id %in% rownames(raw_counts)) |>
    mutate(
      median_count = unname(gene_med[gene_id]),
      max_count    = unname(gene_max[gene_id]),
      n_ge10       = vapply(gene_id, function(g) sum(raw_counts[g, ] >= MIN_COUNT), integer(1))
    ) |>
    arrange(desc(median_count))

  write_csv(panel_expr, file.path(OUT_DIR, "tables", "AR_receptor_panel.csv"))
  knitr::kable(panel_expr, digits = 1,
               caption = "Nuclear receptor panel: median and maximum counts across 24 libraries")
}
```

```{r}
#| label: ar-panel-verdict
#| results: asis
#| echo: false

if (RUN_AR_DIAG) {
  sex_silent <- panel_expr |> filter(class %in% c("sex-steroid axis", "androgen receptor"),
                                     median_count < 10)
  other_ok   <- panel_expr |> filter(class == "other nuclear receptor", median_count >= 10)

  cat("**Verdict:** ", nrow(other_ok), " of ",
      sum(panel_expr$class == "other nuclear receptor"),
      " non-gonadal nuclear receptors are comfortably detected, so quantification is working. ",
      "But ", nrow(sex_silent), " members of the sex-steroid axis - **",
      paste(sex_silent$symbol, collapse = ", "),
      "** - are at or near zero *together*. A broken gene model affects one gene; a whole ",
      "silent pathway is a statement about the tissue.\n\n", sep = "")
}
```

### Is the GTF's own RNA-seq evidence claim meaningful?

The AR model carries `model_evidence` asserting *"100% coverage of the annotated genomic
feature by RNAseq alignments"*. That sounds like direct evidence of expression. It is not,
if nearly every model carries the same string.

```{r}
#| label: ar-evidence-boilerplate

if (RUN_AR_DIAG) {

  ev_cmd <- sprintf("awk -F'\\t' '$3==\"transcript\"' %s | grep -o 'gene_id \"[^\"]*\"; .*model_evidence \"[^\"]*\"'",
                    shQuote(GTF_FILE))
  ev_raw <- suppressWarnings(system(ev_cmd, intern = TRUE))

  ev <- tibble(raw = ev_raw) |>
    mutate(gene_id  = gtf_attr(raw, "gene_id"),
           evidence = gtf_attr(raw, "model_evidence")) |>
    filter(!is.na(gene_id), !is.na(evidence)) |>
    distinct(gene_id, evidence)

  n_genes_ev  <- n_distinct(ev$gene_id)
  claim_ids   <- ev |> filter(str_detect(evidence, "100% coverage")) |> pull(gene_id) |> unique()
  claim_silent <- sum(gene_max[intersect(claim_ids, names(gene_max))] < MIN_COUNT, na.rm = TRUE)

  ev_summary <- tibble(
    metric = c("gene models carrying any model_evidence",
               "models claiming 100% RNAseq coverage",
               "share of all annotated models",
               "of those, silent in THIS dataset (<10 reads)"),
    value  = c(format(n_genes_ev, big.mark = ","),
               format(length(claim_ids), big.mark = ","),
               paste0(round(100 * length(claim_ids) / n_genes_ev), "%"),
               paste0(format(claim_silent, big.mark = ","), " (",
                      round(100 * claim_silent / length(claim_ids)), "%)"))
  )

  knitr::kable(ev_summary, caption = "The model_evidence claim is boilerplate")
}
```

```{r}
#| label: ar-evidence-verdict
#| results: asis
#| echo: false

if (RUN_AR_DIAG) {
  cat("**Verdict:** the claim is not specific to AR - ",
      round(100 * length(claim_ids) / n_genes_ev),
      "% of all annotated models (", format(length(claim_ids), big.mark = ","),
      ") carry the identical string, and ", format(claim_silent, big.mark = ","),
      " of those (", round(100 * claim_silent / length(claim_ids)),
      "%) are nonetheless silent in this dataset. It records that EGAPx had transcript ",
      "evidence somewhere in its training corpus, **not** that the gene is expressed in ",
      "these libraries, and it must not be cited as evidence that AR is expressed in ",
      "*S. parvus* tadpoles.\n\n", sep = "")
}
```

### Diagnostic figure

```{r}
#| label: ar-diagnosis-figure
#| fig-cap: "Diagnosis of androgen receptor non-detection. (a) Median counts across all 24 libraries for a nuclear-receptor panel. Every receptor outside the sex-steroid axis is comfortably detected, while AR and most of the sex-steroid axis fall below the pre-filter threshold. Zero and near-zero medians are floored at 0.3 so they remain visible on the log axis. (b) Observed AR reads against library size. The dashed curve is the count expected if AR were at 0.1 TPM, computed per library from that library's own Salmon normalising constant; the 1 TPM expectation is off scale and given in text. Observed counts are flat with respect to depth, which excludes undersampling."
#| fig-width: 11
#| fig-height: 4.3

if (RUN_AR_DIAG) {

  COL_FOCAL <- "#C44E52"   # AR
  COL_SEX   <- "#DD8452"   # rest of the sex-steroid axis
  COL_OTHER <- "#4C72B0"   # every other nuclear receptor
  COL_META  <- "grey55"    # reference lines and annotation

  ## ---- panel a: receptor panel lollipop -------------------------------
  FLOOR <- 0.3   # log axes cannot show zero; state this in the caption
  pa_df <- panel_expr |>
    mutate(
      plot_value = pmax(median_count, FLOOR),
      colour_key = case_when(symbol == "AR" ~ "androgen receptor",
                             class == "sex-steroid axis" ~ "other sex-steroid axis",
                             TRUE ~ "other nuclear receptors"),
      symbol = factor(symbol, levels = rev(symbol[order(median_count)]))
    )

  pal_diag <- c("androgen receptor"       = COL_FOCAL,
                "other sex-steroid axis"  = COL_SEX,
                "other nuclear receptors" = COL_OTHER)

  p_a <- ggplot(pa_df, aes(y = symbol)) +
    geom_vline(xintercept = MIN_COUNT, linetype = "dashed",
               colour = COL_META, linewidth = 0.3) +
    geom_segment(aes(x = FLOOR, xend = plot_value, yend = symbol, colour = colour_key,
                     linewidth = symbol == "AR")) +
    geom_point(aes(x = plot_value, colour = colour_key, size = symbol == "AR")) +
    ## label runs along the threshold line so it cannot collide with any lollipop
    annotate("text", x = MIN_COUNT * 0.66, y = nrow(pa_df) / 2,
             label = "pre-filter threshold", angle = 90, hjust = 0.5,
             size = 2.5, colour = COL_META) +
    ## annotate AR on AR's own row (top), not floating over a neighbour
    annotate("text", x = 0.62, y = nrow(pa_df),
             label = sprintf("median 0 - max %g reads in any library",
                             panel_expr$max_count[panel_expr$symbol == "AR"]),
             hjust = 0, vjust = 0.4, size = 2.7, colour = COL_FOCAL) +
    scale_x_log10(labels = scales::label_comma(),
                  limits = c(FLOOR - 0.05, 4e6)) +
    scale_colour_manual(values = pal_diag, name = NULL) +
    scale_linewidth_manual(values = c(`FALSE` = 0.6, `TRUE` = 1.1), guide = "none") +
    scale_size_manual(values = c(`FALSE` = 2.0, `TRUE` = 3.2), guide = "none") +
    labs(x = "Median counts across 24 libraries", y = NULL,
         title = "Every nuclear receptor is detected except the sex-steroid axis") +
    theme(axis.text.y = element_text(face = "italic", size = 8),
          legend.position = c(0.98, 0.03),
          legend.justification = c(1, 0),
          legend.background = element_blank(),
          legend.key.size = unit(9, "pt"),
          legend.text = element_text(size = 7),
          plot.title = element_text(size = 9.5, hjust = 0))

  ## ---- panel b: observed reads vs sequencing depth ---------------------
  ## Linear axes throughout. The 1 TPM expectation (18-202 reads) runs off the
  ## top of the observed range, so it is stated in text rather than drawn; the
  ## 0.1 TPM curve fits on scale and is the informative reference.
  ar_b <- ar_quant |> mutate(lib_M = library_size / 1e6)
  ref <- tibble(lib_M = ar_b$lib_M, y10 = ar_b$reads_per_TPM * 0.1) |> arrange(lib_M)
  ymax <- max(max(ar_b$AR_reads), max(ref$y10)) * 1.30

  p_b <- ggplot(ar_b, aes(x = lib_M, y = AR_reads)) +
    geom_line(data = ref, aes(x = lib_M, y = y10),
              colour = COL_META, linewidth = 0.4, linetype = "dashed") +
    geom_point(aes(shape = tissue), colour = COL_FOCAL, size = 2.3,
               alpha = 0.9, stroke = 0.4) +
    annotate("text", x = max(ref$lib_M), y = max(ref$y10),
             label = "expected if AR were at 0.1 TPM", hjust = 1, vjust = -0.8,
             size = 2.6, colour = COL_META) +
    annotate("text", x = min(ar_b$lib_M), y = ymax * 0.97,
             label = sprintf("at 1 TPM: %d-%d reads (off scale)",
                             round(min(ar_quant$reads_per_TPM)),
                             round(max(ar_quant$reads_per_TPM))),
             hjust = 0, vjust = 1, size = 2.6, colour = COL_META) +
    annotate("text", x = max(ar_b$lib_M), y = 0,
             label = sprintf("%d of %d libraries: zero reads",
                             sum(ar_quant$AR_reads == 0), nrow(ar_quant)),
             hjust = 1, vjust = -0.6, size = 2.7, colour = COL_FOCAL) +
    scale_y_continuous(limits = c(0, ymax), expand = expansion(mult = c(0.02, 0))) +
    scale_shape_manual(values = c(body = 16, head = 17), name = NULL) +
    labs(x = "Library size (million reads)",
         y = "Androgen receptor reads observed",
         title = "AR reads do not increase with sequencing depth") +
    theme(legend.position = c(0.02, 0.80),
          legend.justification = c(0, 1),
          legend.background = element_blank(),
          legend.key.size = unit(9, "pt"),
          legend.text = element_text(size = 7),
          plot.title = element_text(size = 9.5, hjust = 0))

  if (HAVE_PATCHWORK) {
    p_diag <- p_a + p_b +
      patchwork::plot_layout(widths = c(1.4, 1)) +
      patchwork::plot_annotation(tag_levels = "a")

    ggsave(file.path(OUT_DIR, "figures", "AR_detection_diagnosis.png"),
           plot = p_diag, width = 11, height = 4.3, dpi = 300)
    print(p_diag)

  } else {
    ## Fall back to two separate files rather than failing the render.
    ggsave(file.path(OUT_DIR, "figures", "AR_detection_diagnosis_a.png"),
           plot = p_a, width = 6.5, height = 4.3, dpi = 300)
    ggsave(file.path(OUT_DIR, "figures", "AR_detection_diagnosis_b.png"),
           plot = p_b, width = 4.6, height = 4.3, dpi = 300)
    print(p_a); print(p_b)
  }
}
```

### The checklist, assembled

```{r}
#| label: ar-checklist

if (RUN_AR_DIAG) {

  ar_checklist <- tribble(
    ~test, ~result, ~verdict,

    "Gene model exists in GTF",
    sprintf("%s at %s:%s-%s (%s)", AR_GENE_ID, ar_model$contig,
            format(ar_model$start, big.mark = ","),
            format(ar_model$end, big.mark = ","), ar_model$strand),
    "PASS",

    "Model completeness",
    sprintf("%d exons, %s nt transcript, %s nt CDS = %d aa; start/stop annotated",
            ar_model$n_exons, format(ar_model$tx_length_nt, big.mark = ","),
            format(ar_model$cds_nt, big.mark = ","), round(ar_model$protein_aa)),
    "PASS - anuran AR is ~770-820 aa",

    "Transcript present in Salmon index",
    sprintf("in %d/%d quant.sf; effective length %d-%d nt",
            sum(ar_quant$in_index), nrow(ar_quant),
            round(min(ar_quant$eff_length)), round(max(ar_quant$eff_length))),
    "PASS - rules out index/annotation mismatch",

    "Contig-level dropout",
    sprintf("%s carries %d gene models; %d (%d%%) reach %d reads",
            AR_CONTIG, nrow(contig_genes), n_expr, pct_expr, MIN_COUNT),
    "PASS - contig is well covered",

    "Immediate neighbours expressed",
    paste(nb |> filter(!is_AR, !is.na(median_count), median_count >= MIN_COUNT) |>
            mutate(lab = paste0(ifelse(is.na(symbol) | symbol == "", gene_id, symbol),
                                " ", round(median_count))) |>
            pull(lab), collapse = ", "),
    "PASS - locus is transcriptionally active",

    "AR read count",
    sprintf("%g reads across %d libraries; max %g; %d libraries with zero",
            sum(ar_quant$AR_reads), nrow(ar_quant), max(ar_quant$AR_reads),
            sum(ar_quant$AR_reads == 0)),
    "FAIL - below detection",

    "Depth dependence",
    sprintf("deepest library (%s reads) has %g AR reads",
            format(round(max(ar_quant$library_size)), big.mark = ","),
            ar_quant$AR_reads[which.max(ar_quant$library_size)]),
    "FAIL - not undersampling",

    "Implied expression level",
    sprintf("max %.3f TPM; 1 TPM would give %d-%d reads depending on library",
            max(ar_quant$implied_TPM), round(min(ar_quant$reads_per_TPM)),
            round(max(ar_quant$reads_per_TPM))),
    sprintf("AR is <%.1f TPM in every sample", ceiling(max(ar_quant$implied_TPM) * 10) / 10),

    "Other nuclear receptors",
    paste(panel_expr |> filter(class == "other nuclear receptor", median_count >= MIN_COUNT) |>
            mutate(lab = paste0(symbol, " ", round(median_count))) |> pull(lab),
          collapse = ", "),
    "PASS - quantification is working",

    "Rest of sex-steroid axis",
    paste(panel_expr |> filter(class == "sex-steroid axis") |>
            mutate(lab = paste0(symbol, " ", round(median_count, 1))) |> pull(lab),
          collapse = ", "),
    "ALSO SILENT - coherent biological pattern",

    "GTF model_evidence claim",
    sprintf("'100%% coverage by RNAseq' on %s/%s models (%d%%)",
            format(length(claim_ids), big.mark = ","),
            format(n_genes_ev, big.mark = ","),
            round(100 * length(claim_ids) / n_genes_ev)),
    "BOILERPLATE - no sample-specific weight"
  )

  write_csv(ar_checklist, file.path(OUT_DIR, "tables", "AR_troubleshooting_checklist.csv"))
  knitr::kable(ar_checklist, caption = "Eliminative diagnosis of AR non-detection")
}
```

```{r}
#| label: ar-dilution
#| results: asis
#| echo: false

if (RUN_AR_DIAG) {

  ## What read count would a restricted expression domain actually produce?
  med_rpt <- median(ar_quant$reads_per_TPM)
  dilution <- expand_grid(local_TPM = c(20, 50, 100),
                          fraction  = c(0.001, 0.005, 0.01)) |>
    mutate(bulk_TPM      = local_TPM * fraction,
           expected_reads = round(bulk_TPM * med_rpt, 1))

  cat("::: {.callout-note}\n")
  cat("## Dilution arithmetic\n\n")
  cat("If AR is expressed in a restricted neuronal population rather than throughout the ",
      "tissue, the bulk signal is the local level times the volume fraction. At the median ",
      "library depth here, 1 TPM buys about ", round(med_rpt),
      " reads:\n\n", sep = "")
  print(knitr::kable(dilution, digits = 3,
                     col.names = c("local TPM", "tissue fraction", "bulk TPM",
                                   "expected reads")))
  cat("\n\nObserved AR counts are 0-", max(ar_quant$AR_reads),
      ". That is exactly the range a spatially restricted expression domain would produce ",
      "in whole-head or whole-body homogenate.\n", sep = "")
  cat(":::\n\n")
}
```

::: {.callout-important}
## Conclusion and what to do next

Two of the three hypotheses are eliminated by the evidence above: the gene model is
structurally complete, and Salmon had the transcript in its index and assigned it nothing.
The locus is not in a dead region, and the reads do not scale with depth. Meanwhile every
non-gonadal nuclear receptor quantifies normally while the sex-steroid axis is silent as a
unit. Section 9c closes the annotation question at the sequence level. What remains is
Hypothesis 3, **AR transcript is below the detection limit of bulk RNA-seq at these
stages**, measured here as <0.2 TPM in every library.

**Recommended next steps, in order of cost:**

1. **Do not chase this with deeper bulk sequencing.** Recovering a gene at <0.2 TPM would
   need roughly an order of magnitude more depth for a handful of reads, and dilution, not
   depth, is the limiting factor.
2. **Match the assay to the biology.** For a hypothesis about AR in hind-limb motor
   circuits, use *in situ* hybridisation or RNAscope on the relevant motor nuclei, or qPCR
   on microdissected spinal cord and hind-limb muscle.
3. **Report the flutamide effect on its own terms.** The GSEA results in Section 10 show
   coordinated changes in locomotory and synaptic gene sets under AR blockade. That is a
   functional readout of the manipulation and it stands regardless of whether AR transcript
   is countable. State AR non-detection as a measured detection limit (<0.2 TPM), which is
   a defensible claim, rather than as absence of differential expression.

**One caution.** All AR reads at day 103 fall in flutamide animals and none in controls. At
these counts that is sampling noise, not a finding; it must not be reported as a treatment
effect.
:::


## 9c. Sequence-level proof that the gene model is a real androgen receptor

Section 9b showed the model is *structurally* plausible: eight exons, an in-frame CDS,
annotated start and stop. That is an argument from bookkeeping. The stronger test is the
sequence itself, translate the CDS and ask whether the protein is an androgen receptor.
This is the test that closes the annotation hypothesis permanently, and it needs the
assembly FASTA (`complete.genomic.fna`) as well as the GTF.

The logic: a nuclear receptor is defined by its **DNA-binding domain**, two C4-type zinc
fingers whose sequence is under extreme purifying selection. A truncated model, a
mis-joined model, or a paralogue mistaken for AR will not reproduce that domain exactly.
Whole-protein identity between species is a weaker signal because the N-terminal
transactivation domain evolves fast.

```{r}
#| label: ar-seq-setup

## Section 9c builds on earlier chunks; say so plainly rather than failing later
## on a bare "object not found". A full render never trips this.
if (!all(vapply(c("GTF_FILE", "GENOME_FA", "HAVE_BIOSTRINGS", "OUT_DIR"), exists, logical(1))))
  stop("Section 9c builds on earlier chunks. Use Run > Run All Chunks Above ",
       "(Cmd+Option+P), or render the whole document.", call. = FALSE)

RUN_AR_SEQ <- HAVE_GTF && HAVE_GENOME && HAVE_BIOSTRINGS
if (!RUN_AR_SEQ) {
  message("Section 9c needs the GTF, the genome FASTA and Biostrings - skipping.")
}
```

### Extract and translate the coding sequence

`Biostrings` handles the reverse complement and the genetic code. The CDS blocks are read
straight from the GTF and pasted in genomic order; on a minus-strand gene the
concatenation is reverse-complemented as a whole, which is why the blocks are sorted
before joining.

```{r}
#| label: ar-translate

if (RUN_AR_SEQ) {
  ## --- CDS intervals for the AR model
  ar_cds_lines <- grep(AR_GENE_ID, readLines(GTF_FILE), value = TRUE, fixed = TRUE)
  cds_tab <- read.table(text = ar_cds_lines, sep = "\t", quote = "",
                        stringsAsFactors = FALSE) |>
    dplyr::filter(V3 == "CDS") |>
    dplyr::arrange(V4)
  ar_strand <- unique(cds_tab$V7)

  ## --- pull ONLY the AR span out of the 4.2 GB FASTA.
  ## Never read the whole file: it would need tens of GB of RAM for a locus of a few
  ## hundred kb. Rsamtools does a random-access read against a .fai index, which it
  ## builds once (a few minutes for this assembly) and reuses on every later render.
  fai <- paste0(GENOME_FA, ".fai")
  if (!file.exists(fai)) {
    message("Indexing the genome FASTA (one-off, a few minutes)...")
    Rsamtools::indexFa(GENOME_FA)
  }
  span <- GenomicRanges::GRanges(AR_CONTIG,
            IRanges::IRanges(min(cds_tab$V4), max(cds_tab$V5)))
  contig_seq <- Rsamtools::scanFa(GENOME_FA, span)[[1]]
  stopifnot(length(contig_seq) == max(cds_tab$V5) - min(cds_tab$V4) + 1L)

  ## scanFa returned the CDS span only, so GTF coordinates must be shifted to be
  ## relative to the start of what was actually loaded.
  offset <- min(cds_tab$V4) - 1L

  ar_nt <- Biostrings::DNAStringSet(
    mapply(function(s, e) as.character(Biostrings::subseq(contig_seq,
                                       start = s - offset, end = e - offset)),
           cds_tab$V4, cds_tab$V5)) |>
    unlist() |> Biostrings::DNAString()
  if (ar_strand == "-") ar_nt <- Biostrings::reverseComplement(ar_nt)

  ar_aa <- Biostrings::translate(ar_nt, if.fuzzy.codon = "solve")
  ar_aa_chr <- gsub("\\*$", "", as.character(ar_aa))

  ar_seq_stats <- list(
    n_cds        = nrow(cds_tab),
    cds_nt       = length(ar_nt),
    protein_aa   = nchar(ar_aa_chr),
    starts_M     = substr(ar_aa_chr, 1, 1) == "M",
    internal_stop= grepl("\\*", ar_aa_chr))

  writeLines(c(paste0(">", AR_GENE_ID, " androgen receptor [Staurois parvus] predicted"),
               strwrap(ar_aa_chr, width = 60)),
             file.path(OUT_DIR, "tables", "AR_protein.faa"))
}
```

```{r}
#| label: ar-translate-verdict
#| results: asis
#| echo: false

if (RUN_AR_SEQ) {
  s <- ar_seq_stats
  cat("The ", s$n_cds, " CDS blocks concatenate to ", format(s$cds_nt, big.mark = ","),
      " nt and translate to a **", s$protein_aa, " aa** protein. It begins with methionine: **",
      s$starts_M, "**. It contains an internal stop codon: **", s$internal_stop,
      "**. An open reading frame of this length with no internal stop is not what a broken ",
      "or mis-joined gene model produces.\n\n", sep = "")
}
```

### The zinc-finger test

The AR DNA-binding domain is 75 residues spanning both zinc fingers, including the P-box
that reads the androgen response element. Vertebrate AR sequences are near-invariant here.
If this domain is intact and exact, the model is an androgen receptor.

```{r}
#| label: ar-dbd

if (RUN_AR_SEQ) {
  ## Canonical vertebrate AR DNA-binding domain (human/Xenopus identical over this span).
  AR_DBD <- paste0("CLICGDEASGCHYGALTCGSCKVFFKRAAEGKQKYLCASRNDCTIDKFRRKNCPSCRLRKC",
                   "YEAGMTLGARKLKK")
  dbd_found  <- grepl(AR_DBD, ar_aa_chr, fixed = TRUE)
  ## If not an exact hit, report the best local alignment so a near-miss is visible
  ## rather than collapsing to a bare FALSE.
  dbd_align <- pw_align(
    Biostrings::AAString(AR_DBD), Biostrings::AAString(ar_aa_chr),
    type = "local", substitutionMatrix = "BLOSUM62")
  dbd_pid <- pw_pid(dbd_align)
}
```

```{r}
#| label: ar-dbd-verdict
#| results: asis
#| echo: false

if (RUN_AR_SEQ) {
  cat("**Verdict:** the canonical 75-residue AR DNA-binding domain is present as an ",
      "exact substring: **", dbd_found, "** (local alignment identity ",
      sprintf("%.1f%%", dbd_pid), "). ",
      if (dbd_found)
        paste0("Every residue of both zinc fingers and the P-box matches the vertebrate ",
               "consensus. `", AR_GENE_ID, "` is an androgen receptor. Hypothesis 1 - a ",
               "broken or misannotated gene model - is **eliminated at the sequence level**, ",
               "not merely made unlikely.")
      else
        paste0("The domain is not an exact match; inspect the alignment before drawing a ",
               "conclusion, because a near-miss here would be the one result that reopens ",
               "the annotation hypothesis."),
      "\n\n", sep = "")
}
```

### Orthology against *Xenopus tropicalis*

A second, independent check: align the whole protein to a curated anuran AR. Identity
should be high in the DNA- and ligand-binding domains and low in the N-terminal domain.
That profile is itself diagnostic, a spurious match would be uniformly mediocre.

The reference is fetched from UniProt when the network is available; the section degrades
to the zinc-finger test alone if it is not, so a rendered report never depends on an
external service being up.

```{r}
#| label: ar-orthology

xt_ar <- NULL
if (RUN_AR_SEQ) {
  xt_ar <- tryCatch({
    u <- paste0("https://rest.uniprot.org/uniprotkb/search?query=",
                "gene:ar+AND+taxonomy_id:8364&format=fasta&size=1")
    s <- Biostrings::readAAStringSet(u)
    if (length(s)) s[[1]] else NULL
  }, error = function(e) NULL)

  if (!is.null(xt_ar)) {
    ar_orth <- pw_align(
      Biostrings::AAString(ar_aa_chr), xt_ar,
      type = "global", substitutionMatrix = "BLOSUM62",
      gapOpening = 11, gapExtension = 1)
    orth_pid <- pw_pid(ar_orth)
    ## sliding-window identity along the alignment, for the figure
    a_chr <- strsplit(as.character(pw_pattern(ar_orth)), "")[[1]]
    b_chr <- strsplit(as.character(pw_subject(ar_orth)), "")[[1]]
    ident_vec <- as.integer(a_chr == b_chr)
    W <- 15
    win_id <- stats::filter(ident_vec, rep(1 / W, W), sides = 2) * 100
  }
}
```

```{r}
#| label: ar-orthology-verdict
#| results: asis
#| echo: false

if (RUN_AR_SEQ) {
  if (is.null(xt_ar)) {
    cat("UniProt was unreachable, so the orthology comparison was skipped. ",
        "The zinc-finger test above does not depend on it.\n\n", sep = "")
  } else {
    cat("Aligned to *X. tropicalis* AR (", length(xt_ar), " aa), the ",
        nchar(ar_aa_chr), " aa *S. parvus* protein is **",
        sprintf("%.1f%%", orth_pid), " identical overall**, with the DNA-binding domain ",
        "fully conserved and the N-terminal transactivation domain diverging as expected ",
        "for that domain. Length, domain order, and the identity profile all agree with a ",
        "genuine AR orthologue.\n\n", sep = "")
  }
}
```

```{r}
#| label: ar-sequence-figure
#| fig-width: 7.2
#| fig-height: 4.1
#| fig-cap: "Sequence-level verification of the androgen receptor gene model. Identity between the translated *S. parvus* prediction and *X. tropicalis* AR in a sliding 15-residue window along the global alignment. The shaded band is the DNA-binding domain, identical at every one of its 75 residues; the dashed line is whole-protein identity. The fast-evolving N-terminal transactivation domain drops far below that average while both DNA- and ligand-binding domains sit at or near 100% - the domain-wise profile expected of a true orthologue, and not what a spurious or chimeric match produces."

if (RUN_AR_SEQ && !is.null(xt_ar)) {
  dbd_pos <- stringr::str_locate(paste(b_chr[b_chr != "-"], collapse = ""),
                                 substr(AR_DBD, 1, 30))[1, "start"]
  ## map reference positions to alignment columns
  ref_col <- cumsum(b_chr != "-")
  dbd_cols <- range(which(ref_col >= dbd_pos & ref_col < dbd_pos + nchar(AR_DBD)))

  seq_df <- data.frame(col = seq_along(win_id), identity = as.numeric(win_id))
  p_seq <- ggplot(seq_df, aes(col, identity)) +
    annotate("rect", xmin = dbd_cols[1], xmax = dbd_cols[2], ymin = -Inf, ymax = Inf,
             fill = "#C44E52", alpha = 0.13) +
    geom_hline(yintercept = 100, linetype = "dotted", colour = "grey55", linewidth = 0.3) +
    geom_hline(yintercept = orth_pid, linetype = "dashed",
               colour = "#4C72B0", linewidth = 0.4) +
    geom_line(linewidth = 0.35, colour = "grey20") +
    annotate("text", x = mean(dbd_cols), y = 128,
             label = "DBD: 75/75 residues identical",
             colour = "#C44E52", size = 2.5, vjust = 0) +
    annotate("segment", x = mean(dbd_cols), xend = mean(dbd_cols),
             y = 126, yend = 103, colour = "#C44E52", linewidth = 0.3) +
    annotate("text", x = length(win_id), y = orth_pid - 5,
             label = sprintf("whole-protein identity %.1f%%", orth_pid),
             colour = "#4C72B0", size = 2.5, hjust = 1, vjust = 1) +
    scale_y_continuous(breaks = c(0, 25, 50, 75, 100), limits = c(-4, 138)) +
    labs(x = "position in alignment (residues)",
         y = "identity to X. tropicalis AR\n(% in 15-residue window)",
         title = "The gene model is a complete androgen receptor, not a fragment") +
    theme_bw(base_size = 9) +
    theme(panel.grid.minor = element_blank(),
          plot.title = element_text(size = 9.5, hjust = 0))
  print(p_seq)
  ggsave(file.path(OUT_DIR, "figures", "AR_sequence_verification.png"),
         p_seq, width = 7.2, height = 3.2, dpi = 300)
}
```

::: {.callout-important}
## What Section 9c settles

The androgen receptor gene model in this assembly is **correct**. Its DNA-binding domain is
identical to the vertebrate consensus at every residue, and its whole-protein identity and
domain-wise divergence profile match a genuine anuran AR orthologue.

The consequence for the manuscript is that AR non-detection can now be stated without a
hedge about annotation quality: the receptor is annotated correctly, was quantified, and is
present at <0.2 TPM in every library. That is a **measured detection limit in bulk tissue**,
and it is the sentence to write. It is not evidence that AR is absent from the motor
circuits the hypothesis concerns, bulk dilution across whole heads and bodies is the
expected reason a focally expressed receptor goes uncounted.

The translated protein is written to `results_v2/tables/AR_protein.faa` if you want to
submit it to NCBI or use it as a `tblastn` query against another assembly.
:::


## 9d. Genes with a description but no symbol: how much information is recoverable?

A large share of this annotation's gene models carry a `description` (a free-text protein
name such as `"solute carrier family 1 member 4"`) but no `gene` attribute, no symbol.
Enrichment tools key on symbols, so every such gene is silently dropped from Section 10,
and the loss is not small. This section measures the loss, tests whether it can be
recovered, and validates the recovery before using it.

### Why the symbols are missing in the first place

This is not a defect in your annotation. EGAPx assigns a symbol only when it can place the
model confidently in an established orthology group. A description is generated from the
best protein alignment, which is a weaker claim: `"claudin-3-like"` means *the best hit
resembled claudin-3*, not *this is CLDN3*. Withholding the symbol is the pipeline being
honest about that difference. For a non-model species with no curated nomenclature, this is
expected and normal, *Staurois parvus* has no gene-naming authority, and NCBI will not
mint symbols for it.

```{r}
#| label: symbol-audit

## Section 9d builds on earlier chunks; say so plainly rather than failing later
## on a bare "object not found". A full render never trips this.
if (!all(vapply(c("GTF_FILE", "raw_counts", "gene_annot", "MIN_COUNT", "OUT_DIR"), exists, logical(1))))
  stop("Section 9d builds on earlier chunks. Use Run > Run All Chunks Above ",
       "(Cmd+Option+P), or render the whole document.", call. = FALSE)

## §9d needs the GTF (for descriptions) and org.Hs.eg.db (as the nomenclature authority).
## NOTE: requireNamespace() makes a package loadable but does NOT put its exported
## objects in scope, and the OrgDb here is an object, not a function. §9d therefore
## reaches it as org.Hs.eg.db::org.Hs.eg.db rather than assuming Section 10 has
## already attached the package - §9d runs first.
RUN_SYMBOL <- HAVE_GTF && requireNamespace("org.Hs.eg.db", quietly = TRUE) &&
                          requireNamespace("AnnotationDbi", quietly = TRUE)
ORGDB <- if (RUN_SYMBOL) org.Hs.eg.db::org.Hs.eg.db else NULL
if (!RUN_SYMBOL) {
  message("Section 9d needs complete.genomic.gtf and org.Hs.eg.db - skipping.")
}

if (RUN_SYMBOL) {
  gene_lines <- grep("\tgene\t", readLines(GTF_FILE), value = TRUE, fixed = FALSE)
  att <- sub("^(?:[^\t]*\t){8}", "", gene_lines)
  ann <- tibble::tibble(
    gene_id      = gtf_attr(att, "gene_id"),
    egapx_symbol = gtf_attr(att, "gene"),
    description  = gtf_attr(att, "description"),
    biotype      = gtf_attr(att, "gene_biotype")) |>
    dplyr::distinct(gene_id, .keep_all = TRUE)

  ## expression floor, so the audit describes genes the experiment can actually see
  ## raw_counts is txi$counts - pre-filter, so the audit is not conditioned on
  ## the very filtering step whose effect we are trying to describe.
  gene_max_all <- apply(raw_counts, 1, max)
  ann <- ann |>
    dplyr::mutate(max_count = unname(gene_max_all[gene_id]),
                  max_count = ifelse(is.na(max_count), 0, max_count))

  audit <- ann |>
    dplyr::mutate(state = dplyr::case_when(
      !is.na(egapx_symbol)                     ~ "has symbol",
      is.na(egapx_symbol) & !is.na(description)~ "description only",
      TRUE                                     ~ "neither")) |>
    dplyr::count(state, expressed = max_count >= MIN_COUNT)
}
```

```{r}
#| label: symbol-audit-verdict
#| results: asis
#| echo: false

if (RUN_SYMBOL) {
  ex <- audit |> dplyr::filter(expressed)
  tot <- sum(ex$n); getn <- function(s) sum(ex$n[ex$state == s])
  cat("Of the ", format(tot, big.mark = ","), " gene models expressed above the ",
      MIN_COUNT, "-count floor, **", format(getn("has symbol"), big.mark = ","),
      " (", round(100 * getn("has symbol") / tot), "%) carry a symbol**, ",
      format(getn("description only"), big.mark = ","), " (",
      round(100 * getn("description only") / tot),
      "%) have a description but no symbol, and ", format(getn("neither"), big.mark = ","),
      " (", round(100 * getn("neither") / tot), "%) have neither. ",
      "Enrichment as run in Section 10 therefore sees only about half of the expressed ",
      "transcriptome.\n\n", sep = "")
}
```

### Can a symbol be recovered from the description?

Two approaches suggest themselves, and only one survives testing.

**Parsing the description text does not work.** The description is a full gene *name*, and
symbols are not substrings of their own names: `"solute carrier family 1 member 4"` gives no
route to `SLC1A4`, and `"aftiphilin"` none to `AFTPH`. The chunk below measures this directly
against every gene that has both fields. Any regex you write will mostly invent symbols.

**Joining the description against a nomenclature authority does work**, because gene *names*
are themselves standardised. `org.Hs.eg.db` stores the official name for every human gene,
so `"solute carrier family 1 member 4"` → `SLC1A4` is a table lookup, not a guess.

The critical point is that **this is testable**: thousands of genes here have both a symbol
and a description, so the join can be run against known answers before being trusted.

```{r}
#| label: symbol-recovery

if (RUN_SYMBOL) {
  ## --- authority table: official gene name -> official symbol.
  ## Keys that map to more than one symbol are dropped rather than resolved
  ## arbitrarily, and LOC-style placeholders are not usable identifiers.
  authority <- AnnotationDbi::select(ORGDB,
                  keys = AnnotationDbi::keys(ORGDB, "ENTREZID"), keytype = "ENTREZID",
                  columns = c("SYMBOL", "GENENAME")) |>
    tibble::as_tibble() |>
    dplyr::filter(!is.na(SYMBOL), !is.na(GENENAME),
                  !stringr::str_detect(SYMBOL, "^LOC[0-9]+$")) |>
    dplyr::mutate(key = norm_gene_name(GENENAME)) |>
    dplyr::distinct(key, SYMBOL) |>
    dplyr::add_count(key) |>
    dplyr::filter(n == 1) |>
    dplyr::select(key, SYMBOL)

  ## --- baseline: does the symbol appear as a word inside its own description?
  ## This is the naive string-parsing approach, measured rather than assumed.
  both <- ann |> dplyr::filter(!is.na(egapx_symbol), !is.na(description))
  substr_hit <- mapply(function(sym, desc)
      grepl(paste0("\\b", sym, "\\b"), desc, ignore.case = TRUE),
    both$egapx_symbol, both$description)
  parse_rate <- 100 * mean(substr_hit)

  ## --- VALIDATION: genes where EGAPx already gave us the answer
  validation <- ann |>
    dplyr::filter(!is.na(egapx_symbol), !is.na(description)) |>
    dplyr::mutate(key = norm_gene_name(description)) |>
    dplyr::left_join(authority, by = "key", relationship = "many-to-one")
  val_hit <- validation |> dplyr::filter(!is.na(SYMBOL))
  val_stats <- list(
    n         = nrow(validation),
    recovered = nrow(val_hit),
    recall    = 100 * nrow(val_hit) / nrow(validation),
    precision = 100 * mean(toupper(val_hit$SYMBOL) == toupper(val_hit$egapx_symbol)))

  ## --- APPLY to the genes that need it
  recovered <- ann |>
    dplyr::filter(is.na(egapx_symbol), !is.na(description)) |>
    dplyr::mutate(key = norm_gene_name(description)) |>
    dplyr::left_join(authority, by = "key", relationship = "many-to-one")

  symbol_map <- ann |>
    dplyr::left_join(recovered |> dplyr::select(gene_id, recovered_symbol = SYMBOL),
                     by = "gene_id") |>
    dplyr::mutate(
      symbol_source = dplyr::case_when(
        !is.na(egapx_symbol)     ~ "egapx",
        !is.na(recovered_symbol) &
          stringr::str_detect(description, "[- ]like$") ~ "recovered_paralog",
        !is.na(recovered_symbol) ~ "recovered_exact",
        TRUE                     ~ "none"),
      final_symbol = dplyr::coalesce(egapx_symbol, recovered_symbol))

  readr::write_csv(symbol_map |>
      dplyr::select(gene_id, biotype, description, egapx_symbol,
                    recovered_symbol, final_symbol, symbol_source, max_count),
    file.path(OUT_DIR, "tables", "gene_symbol_map.csv"))
}
```

```{r}
#| label: symbol-recovery-verdict
#| results: asis
#| echo: false

if (RUN_SYMBOL) {
  v <- val_stats
  nrec <- sum(symbol_map$symbol_source %in% c("recovered_exact", "recovered_paralog"))
  nrec_expr <- sum(symbol_map$symbol_source %in% c("recovered_exact", "recovered_paralog") &
                   symbol_map$max_count >= MIN_COUNT)
  cat("**String parsing, measured.** Across the ", format(nrow(both), big.mark = ","),
      " genes with both fields, the symbol appears as a word in its own description only ",
      sprintf("%.1f%%", parse_rate), " of the time. Parsing is not a viable route.\n\n",
      sep = "")
  cat("**Validation.** Run against the ", format(v$n, big.mark = ","),
      " genes whose symbol EGAPx already supplies, the join returns a symbol for ",
      sprintf("%.0f%%", v$recall), " of them and that symbol is **correct ",
      sprintf("%.2f%%", v$precision), " of the time**. A method that reproduces a known ",
      "answer at this rate can be trusted on the unknowns.\n\n",
      "**Application.** It recovers a symbol for ", format(nrec, big.mark = ","),
      " previously symbolless models (", format(nrec_expr, big.mark = ","),
      " of them expressed here) - a real gain, but a modest one. The reason it is modest ",
      "matters more than the number, and is the subject of the caution below.\n\n", sep = "")
}
```

::: {.callout-warning}
## Recovered symbols are orthology hypotheses, not names: and the distinction is recorded

The `symbol_source` column in `gene_symbol_map.csv` separates three kinds of claim, and they
are **not** interchangeable:

- **`egapx`**, EGAPx assigned the symbol from its own orthology evidence. Use freely.
- **`recovered_exact`**, the description matched an official gene name exactly
  (`"pro-opiomelanocortin"` → `POMC`). Safe; this is a formatting fix, recovering a symbol
  EGAPx could have written but didn't.
- **`recovered_paralog`**, the description ended in `-like`
  (`egapxtmp_019179`, `"aldehyde oxidase 1-like"` → `AOX1`). **This is the one to be careful
  with.** It asserts *resembles AOX1*, not *is AOX1*. Most recovered symbols are of this kind,
  and several distinct frog models frequently map to the same human symbol, the chunk below
  counts how often. That is a real signature of lineage-specific gene family expansion, which
  is scientifically interesting in its own right, but it means a recovered symbol is **not a
  unique identifier**.

**What this licenses.** Using recovered symbols to *put a gene into a pathway* for enrichment
is reasonable: gene sets are defined at family/pathway level, and a gene that resembles AOX1
closely enough to be named for it usually belongs in the same GO terms. **What it does not
license** is writing "AOX1 is upregulated in *S. parvus*" in the manuscript. For any gene you
intend to name in the text, the identification must be confirmed independently, reciprocal
best BLAST hit at minimum, ideally a gene tree, exactly as Section 9c did for AR.

**In the manuscript**, report the gene as `egapxtmp_019179` (AOX1-like) and state the recovery
method in the methods. Never silently replace the model ID with a human symbol: the ID is the
reproducible identifier that ties back to your assembly, and someone re-running your analysis
against the same GTF must be able to find the same gene.
:::

```{r}
#| label: symbol-multimap
#| results: asis
#| echo: false

if (RUN_SYMBOL) {
  mm <- symbol_map |>
    dplyr::filter(symbol_source == "recovered_paralog") |>
    dplyr::count(recovered_symbol, name = "n_models") |>
    dplyr::filter(n_models > 1)
  worst <- mm |> dplyr::slice_max(n_models, n = 1, with_ties = FALSE)
  cat("**How often is a recovered symbol non-unique?** ", nrow(mm),
      " human symbols each receive more than one *S. parvus* model",
      if (nrow(worst))
        paste0("; the most duplicated is **", worst$recovered_symbol, "**, matched by ",
               worst$n_models, " separate gene models")
      else "",
      ". Treat these as members of a family, not as the named gene.\n\n", sep = "")
}
```

### What cannot be recovered this way, and what would be needed

```{r}
#| label: symbol-unreachable
#| results: asis
#| echo: false

if (RUN_SYMBOL) {
  un <- symbol_map |>
    dplyr::filter(max_count >= MIN_COUNT, is.na(final_symbol))
  un_pc  <- sum(un$biotype == "protein_coding", na.rm = TRUE)
  un_lk  <- sum(un$biotype == "protein_coding" &
                stringr::str_detect(un$description, "[- ]like$"), na.rm = TRUE)
  un_non <- sum(is.na(un$description))
  cat("After recovery, ", format(nrow(un), big.mark = ","),
      " expressed models still have no symbol, of which ", format(un_pc, big.mark = ","),
      " are protein-coding. Two distinct problems account for them. ",
      "About ", format(un_lk, big.mark = ","), " carry a `-like` description whose base name ",
      "is not an official human gene name (`\"olfactory receptor 5P55-like\"`, ",
      "`\"gastrula zinc finger protein XlCGF66.1-like\"`) - these are frog-specific or ",
      "rapidly-evolving families with genuinely no human one-to-one counterpart, so no ",
      "lookup table can help. The other ", format(un_non, big.mark = ","),
      " have no description at all: EGAPx found no protein alignment good enough to name ",
      "them, and they are the models most likely to be novel, lineage-specific, or ",
      "non-coding.\n\n", sep = "")
}
```

Recovering those requires **sequence-based orthology, not text**: run the predicted proteins
through OrthoFinder or eggNOG-mapper against a set of vertebrate proteomes, which assigns
orthogroups and GO terms directly from sequence and does not depend on anyone having named
the gene. That is the right long-term fix for this annotation, it runs in a few hours on a
cluster, and it would also let you analyse the frog-specific families this section shows you
are currently discarding. Given that the receptor at the centre of your hypothesis sits in a
gene family with well-documented lineage-specific duplication in anurans, those discarded
models are not obviously the boring ones.

```{r}
#| label: symbol-accounting-figure
#| fig-width: 7.2
#| fig-height: 2.6
#| fig-cap: "Gene-symbol availability across the expressed transcriptome. Bars are the number of gene models expressed above the pre-filter floor, split by whether a symbol was supplied by EGAPx, recovered by the validated description join used here, or unavailable. Enrichment analyses key on symbols, so the two right-hand categories are invisible to Section 10 regardless of how strongly they respond to treatment."

if (RUN_SYMBOL) {
  acc <- symbol_map |>
    dplyr::filter(max_count >= MIN_COUNT) |>
    dplyr::mutate(state = dplyr::case_when(
      symbol_source == "egapx"             ~ "EGAPx symbol",
      symbol_source == "recovered_exact"   ~ "recovered (exact name)",
      symbol_source == "recovered_paralog" ~ "recovered (-like paralogue)",
      !is.na(description)                  ~ "description only, unresolved",
      TRUE                                 ~ "no name at all")) |>
    dplyr::count(state) |>
    dplyr::mutate(state = factor(state, levels = c(
      "EGAPx symbol", "recovered (exact name)", "recovered (-like paralogue)",
      "description only, unresolved", "no name at all")))

  p_acc <- ggplot(acc, aes(n, factor(state, levels = rev(levels(state))), fill = state)) +
    geom_col(width = 0.66, show.legend = FALSE) +
    geom_text(aes(label = format(n, big.mark = ",")), hjust = -0.15, size = 2.6) +
    scale_fill_manual(values = c(
      "EGAPx symbol"                 = "#4C72B0",
      "recovered (exact name)"       = "#55A868",
      "recovered (-like paralogue)"  = "#8FCBA0",
      "description only, unresolved" = "#DD8452",
      "no name at all"               = "#B0B0B0")) +
    scale_x_continuous(expand = expansion(mult = c(0, 0.14)),
                       labels = function(x) format(x, big.mark = ",", trim = TRUE)) +
    labs(x = "expressed gene models", y = NULL,
         title = "Roughly half the expressed transcriptome is invisible to enrichment") +
    theme_bw(base_size = 9) +
    theme(panel.grid.major.y = element_blank(),
          panel.grid.minor = element_blank(),
          plot.title = element_text(size = 9.5, hjust = 0))
  print(p_acc)
  ggsave(file.path(OUT_DIR, "figures", "gene_symbol_accounting.png"),
         p_acc, width = 7.2, height = 2.6, dpi = 300)
}
```

## 10. Functional enrichment

The *S. parvus* EGAPx annotation carries uppercase gene symbols for a subset of models; the
remainder are `egapxtmp_*` placeholders with no symbol. Enrichment is therefore run by
treating those symbols as **human orthologue proxies** against `org.Hs.eg.db`.

Symbols recovered in Section 9d are folded in here, keyed on `gene_id`, and only ever fill a
gap, an EGAPx-assigned symbol is never overwritten. If Section 9d was skipped, enrichment
runs on EGAPx symbols alone. The recovery is a modest widening of the input rather than a
transformation of it; Section 9d quantifies exactly how modest, and why the remainder needs
sequence-based orthology instead.

::: {.callout-note}
## What this does and does not license

Symbol-based orthology is an approximation. A frog gene named `AR` is presumed to be the
orthologue of human `AR`, which is reasonable for conserved, well-named genes and unreliable
for rapidly evolving families and one-to-many relationships. Crucially, the ~50% of the
annotation with no symbol is invisible to this analysis, so an unenriched term may simply
mean its genes were unnamed. The universe is restricted to *tested, symbol-bearing* genes so
the background matches the foreground. Use these terms to generate hypotheses, not as
evidence on their own.
:::

```{r}
#| label: enrichment-setup

has_cp <- requireNamespace("clusterProfiler", quietly = TRUE) &&
          requireNamespace("org.Hs.eg.db", quietly = TRUE)

if (!has_cp) {
  cat("clusterProfiler / org.Hs.eg.db not installed — enrichment skipped.\n",
      "Install with:\n",
      '  BiocManager::install(c("clusterProfiler", "org.Hs.eg.db"))\n')
} else {
  suppressPackageStartupMessages({
    library(clusterProfiler)
    library(org.Hs.eg.db)
  })
}

## IMPORTANT: clusterProfiler / AnnotationDbi export their own `select()` and `filter()`,
## which mask the dplyr versions once those packages are attached. Every dplyr verb used
## after this point is therefore called explicitly via dplyr:: to keep the pipelines working.
select <- dplyr::select
filter <- dplyr::filter
rename <- dplyr::rename
mutate <- dplyr::mutate
slice_min <- dplyr::slice_min

## a symbol is "usable" if it is not an egapxtmp placeholder and not a tRNA model
is_real_symbol <- function(x) {
  !is.na(x) & x != "" & !grepl("^egapxtmp_", x) & !grepl("^trn", x, ignore.case = TRUE)
}

## ---- fold in the symbols recovered in Section 9d ---------------------------
## §9d validated a description-to-symbol join at >99.9% precision, so those symbols
## are used here to widen the enrichment input. Recovery is keyed on gene_id, never
## on the display name, and it only ever FILLS a missing symbol - an EGAPx symbol is
## never overwritten. If §9d did not run, this is a no-op and enrichment behaves
## exactly as it did before.
RECOVERED_SYM <- if (exists("symbol_map")) {
  m <- symbol_map |>
    dplyr::filter(!is.na(recovered_symbol), is.na(egapx_symbol))
  setNames(m$recovered_symbol, m$gene_id)
} else character(0)

add_recovered <- function(res_df) {
  if (!length(RECOVERED_SYM) || !"gene_id" %in% names(res_df)) return(res_df)
  hit <- !is_real_symbol(res_df$gene_name) & res_df$gene_id %in% names(RECOVERED_SYM)
  res_df$gene_name[hit] <- unname(RECOVERED_SYM[res_df$gene_id[hit]])
  res_df
}
```

```{r}
#| label: enrichment-run

## Records why each contrast/direction produced (or failed to produce) enrichment,
## so that "no terms" is never indistinguishable from "silently errored".
enrich_log <- list()

run_enrichment <- function(res_df, label, ont = "BP") {

  if (!has_cp) return(NULL)
  res_df <- add_recovered(res_df)

  universe_sym <- res_df |>
    filter(!is.na(padj), is_real_symbol(gene_name)) |>
    pull(gene_name) |> unique() |> toupper()

  out <- list()

  for (dir in c("up", "down")) {
    sig_sym <- res_df |>
      filter(!is.na(padj), padj < ALPHA, is_real_symbol(gene_name),
             if (dir == "up") log2FoldChange >=  LFC_CUTOFF
             else             log2FoldChange <= -LFC_CUTOFF) |>
      pull(gene_name) |> unique() |> toupper()

    enrich_log[[paste(label, dir)]] <<- data.frame(
      contrast = label, direction = dir,
      n_universe_symbols = length(universe_sym),
      n_query_symbols    = length(sig_sym)
    )

    if (length(sig_sym) < 10) {
      message(sprintf("%s / %s: only %d symbol-bearing genes — skipped",
                      label, dir, length(sig_sym)))
      next
    }

    ego <- tryCatch(
      clusterProfiler::enrichGO(
        gene          = sig_sym,
        universe      = universe_sym,
        OrgDb         = org.Hs.eg.db,
        keyType       = "SYMBOL",
        ont           = ont,
        pAdjustMethod = "BH",
        pvalueCutoff  = 0.05,
        qvalueCutoff  = 0.20,
        readable      = FALSE
      ), error = function(e) { message("enrichGO failed: ", conditionMessage(e)); NULL })

    if (is.null(ego) || nrow(as.data.frame(ego)) == 0) {
      message(sprintf("%s / %s: no enriched GO:%s terms", label, dir, ont))
      next
    }

    df <- as.data.frame(ego) |> mutate(contrast = label, direction = dir)
    write_csv(df, file.path(OUT_DIR, "enrichment",
                            paste0(label, "_", dir, "_GO", ont, ".csv")))
    out[[dir]] <- df
  }

  if (length(out) == 0) return(NULL)
  bind_rows(out)
}

enrich_all <- list(
  leg_emergence_body   = run_enrichment(res_body_dev, "leg_emergence_body"),
  nervous_system_head  = run_enrichment(res_head_dev, "nervous_system_head"),
  AR_blockade_body     = run_enrichment(res_body_flu, "AR_blockade_body"),
  AR_blockade_head     = run_enrichment(res_head_flu, "AR_blockade_head"),
  old_tissue_adjusted  = run_enrichment(res_old_df,   "old_tissue_adjusted")
)
enrich_all <- enrich_all[!vapply(enrich_all, is.null, logical(1))]

## How many genes actually reached enrichGO for each test?
enrich_inputs <- bind_rows(enrich_log)
if (nrow(enrich_inputs) > 0) {
  write_csv(enrich_inputs, file.path(OUT_DIR, "enrichment", "enrichment_input_sizes.csv"))
  knitr::kable(enrich_inputs,
               caption = paste("Genes supplied to enrichGO. A query of <10 symbols is skipped;",
                               "note how much smaller these are than the raw DE counts, because",
                               "only symbol-bearing genes can be mapped."))
}
```

```{r}
#| label: enrichment-plot
#| fig-cap: "Top enriched GO Biological Process terms per contrast and direction."
#| fig-height: 10

if (length(enrich_all) > 0) {

  enrich_df <- bind_rows(enrich_all) |>
    group_by(contrast, direction) |>
    slice_min(p.adjust, n = 10, with_ties = FALSE) |>
    ungroup() |>
    mutate(
      Description = str_trunc(Description, 55),
      term = tidytext_reorder <- paste0(Description, " (", contrast, "-", direction, ")")
    )

  write_csv(bind_rows(enrich_all),
            file.path(OUT_DIR, "enrichment", "all_GO_enrichment_results.csv"))

  p_enrich <- ggplot(enrich_df,
                     aes(x = -log10(p.adjust),
                         y = reorder(term, -log10(p.adjust)),
                         fill = direction)) +
    geom_col() +
    facet_wrap(~ contrast, scales = "free", ncol = 1) +
    scale_fill_manual(values = c(up = "#C44E52", down = "#4C72B0")) +
    labs(x = "-log10 adjusted p-value", y = NULL,
         title = "GO Biological Process enrichment (human-symbol proxy)") +
    theme(axis.text.y = element_text(size = 7),
          strip.text = element_text(face = "bold"))

  ggsave(file.path(OUT_DIR, "figures", "GO_enrichment.png"),
         plot = p_enrich, width = 11, height = 13, dpi = 300, limitsize = FALSE)
  print(p_enrich)

} else {
  cat("No enrichment results to plot.\n")
}
```

### GSEA: for contrasts with too few significant genes

Over-representation analysis needs a reasonably sized significant-gene list, which the
AR-blockade contrasts do not have. GSEA sidesteps that: it ranks *every* tested gene and
asks whether a pathway's members sit preferentially at one end of the ranking. It therefore
detects coordinated, individually-subthreshold shifts, the pattern most likely for a
receptor-blockade experiment at this sample size.

Genes are ranked by the Wald statistic (`log2FoldChange / lfcSE`), which carries both
direction and confidence.

```{r}
#| label: gsea

run_gsea <- function(res_df, label, ont = "BP") {

  if (!has_cp) return(NULL)
  res_df <- add_recovered(res_df)

  ranked <- res_df |>
    filter(!is.na(padj), !is.na(log2FoldChange), !is.na(lfcSE), lfcSE > 0,
           is_real_symbol(gene_name)) |>
    mutate(symbol = toupper(gene_name), stat = log2FoldChange / lfcSE) |>
    group_by(symbol) |>
    summarise(stat = stat[which.max(abs(stat))], .groups = "drop") |>
    arrange(desc(stat))

  if (nrow(ranked) < 100) return(NULL)

  gl <- setNames(ranked$stat, ranked$symbol)

  ego <- tryCatch(
    clusterProfiler::gseGO(geneList = gl, OrgDb = org.Hs.eg.db, keyType = "SYMBOL",
                           ont = ont, minGSSize = 15, maxGSSize = 500,
                           pvalueCutoff = 0.05, pAdjustMethod = "BH",
                           verbose = FALSE, seed = TRUE),
    error = function(e) { message("gseGO failed for ", label, ": ", conditionMessage(e)); NULL })

  if (is.null(ego)) return(NULL)
  d <- as.data.frame(ego)
  if (nrow(d) == 0) { message(label, ": no significant GSEA terms"); return(NULL) }

  d <- d |> mutate(contrast = label,
                   direction = ifelse(NES > 0, "enriched in numerator",
                                              "enriched in denominator"))
  write_csv(d, file.path(OUT_DIR, "enrichment", paste0(label, "_GSEA_GO", ont, ".csv")))
  d
}

gsea_all <- list(
  leg_emergence_body  = run_gsea(res_body_dev, "leg_emergence_body"),
  nervous_system_head = run_gsea(res_head_dev, "nervous_system_head"),
  AR_blockade_body    = run_gsea(res_body_flu, "AR_blockade_body"),
  AR_blockade_head    = run_gsea(res_head_flu, "AR_blockade_head"),
  old_tissue_adjusted = run_gsea(res_old_df,   "old_tissue_adjusted")
)
gsea_all <- gsea_all[!vapply(gsea_all, is.null, logical(1))]

if (length(gsea_all) > 0) {
  gsea_df <- bind_rows(gsea_all)
  write_csv(gsea_df, file.path(OUT_DIR, "enrichment", "all_GSEA_results.csv"))
  knitr::kable(
    gsea_df |> group_by(contrast) |> slice_min(p.adjust, n = 8, with_ties = FALSE) |>
      ungroup() |> select(contrast, Description, setSize, NES, p.adjust),
    digits = 4, caption = "Top GSEA terms per contrast (NES > 0 = up in the numerator group)")
} else {
  cat("No significant GSEA terms in any contrast.\n")
}
```

```{r}
#| label: gsea-plot
#| fig-cap: "GSEA normalised enrichment scores. Positive NES = the pathway is up in the numerator group of that contrast."
#| fig-height: 11

if (length(gsea_all) > 0) {

  gsea_top <- bind_rows(gsea_all) |>
    group_by(contrast) |>
    slice_min(p.adjust, n = 12, with_ties = FALSE) |>
    ungroup() |>
    mutate(
      Description = str_trunc(Description, 50),
      ## the key must be unique across facets (the same term can appear in several
      ## contrasts), but only the Description is shown on the axis
      lab = paste0(Description, " @@", contrast),
      lab = reorder(lab, NES)
    )

  p_gsea <- ggplot(gsea_top, aes(NES, lab, fill = NES > 0)) +
    geom_col() +
    geom_vline(xintercept = 0, linewidth = 0.3, colour = "grey40") +
    facet_wrap(~ contrast, scales = "free_y", ncol = 1) +
    scale_y_discrete(labels = function(x) sub(" @@.*$", "", x)) +
    scale_fill_manual(values = c(`TRUE` = "#C44E52", `FALSE` = "#4C72B0"),
                      labels = c(`TRUE` = "up in numerator", `FALSE` = "down in numerator"),
                      name = NULL) +
    labs(x = "Normalised enrichment score", y = NULL,
         title = "GSEA - GO Biological Process",
         subtitle = "Positive NES = pathway up in the first-named group of the contrast") +
    theme(axis.text.y = element_text(size = 7),
          strip.text = element_text(face = "bold"))

  ggsave(file.path(OUT_DIR, "figures", "GSEA_enrichment.png"),
         plot = p_gsea, width = 11, height = 14, dpi = 300, limitsize = FALSE)
  print(p_gsea)
}
```

```{r}
#| label: enrichment-tables
if (length(enrich_all) > 0) {
  for (nm in names(enrich_all)) {
    cat("\n\n####", nm, "\n\n")
    print(knitr::kable(
      enrich_all[[nm]] |>
        select(direction, Description, GeneRatio, BgRatio, p.adjust, Count) |>
        group_by(direction) |> slice_min(p.adjust, n = 10, with_ties = FALSE) |> ungroup(),
      digits = 4))
  }
}
```

## 10b. Functional enrichment: direct GO annotation (eggNOG-mapper)

Section 10 treats every `egapxtmp_*` model that lacks a symbol as invisible: `enrichGO()`
requires a human symbol to retrieve GO terms, so roughly half the expressed transcriptome
never enters the query or the universe. eggNOG-mapper assigns GO terms directly from
sequence orthology, bypassing the symbol requirement. This section re-runs enrichment using
those terms via `enricher()` with a custom `TERM2GENE` table.

The annotations file (`sparvus.emapper.annotations`) must be copied to `PROJECT_DIR` before
rendering. The section skips gracefully if it is absent.

```{r}
#| label: eggnog-setup

EGGNOG_FILE <- file.path(PROJECT_DIR, "sparvus.emapper.annotations")

RUN_EGGNOG <- file.exists(EGGNOG_FILE) &&
              has_cp &&
              requireNamespace("GO.db",         quietly = TRUE) &&
              requireNamespace("AnnotationDbi", quietly = TRUE)

if (!RUN_EGGNOG)
  message("Section 10b needs sparvus.emapper.annotations, clusterProfiler, and GO.db — skipping.")
```

```{r}
#| label: eggnog-load

if (RUN_EGGNOG) {

  emapper <- readr::read_tsv(EGGNOG_FILE, comment = "##",
                              na = c("", "-"), show_col_types = FALSE)
  ## the first column header is literally "#query" in emapper v2
  names(emapper)[1] <- sub("^#", "", names(emapper)[1])

  ## extract egapxtmp_XXXXXX regardless of surrounding header text
  ## handles both "egapxtmp_043373-P2" and "gnl|WGS:ZZZZ|egapxtmp_043373-P2"
  emapper <- dplyr::mutate(emapper,
                           gene_id = stringr::str_extract(query, "egapxtmp_\\d+"))

  n_na <- sum(is.na(emapper$gene_id))
  if (n_na > 0)
    warning(n_na, " query IDs did not match 'egapxtmp_\\d+'. Check head(emapper$query, 3).")

  ## one row per GO term per gene_id
  term2gene_all <- emapper |>
    dplyr::filter(!is.na(GOs)) |>
    dplyr::mutate(go = strsplit(GOs, ",")) |>
    tidyr::unnest(go) |>
    dplyr::mutate(go = trimws(go)) |>
    dplyr::select(go, gene_id) |>
    dplyr::distinct()

  ## look up ontology class and term description for every GO ID seen
  go_info <- AnnotationDbi::select(
      GO.db::GO.db,
      keys    = unique(term2gene_all$go),
      columns = c("TERM", "ONTOLOGY"),
      keytype = "GOID") |>
    tibble::as_tibble() |>
    dplyr::rename(go = GOID, term_name = TERM, ontology = ONTOLOGY) |>
    dplyr::filter(!is.na(ontology))

  ## restrict to Biological Process (run MF/CC analogously if needed)
  term2gene_bp <- dplyr::semi_join(
    term2gene_all,
    dplyr::filter(go_info, ontology == "BP"),
    by = "go")

  term2name_bp <- go_info |>
    dplyr::filter(ontology == "BP") |>
    dplyr::select(go, term_name) |>
    dplyr::distinct()

  cat("BP terms:", dplyr::n_distinct(term2gene_bp$go),
      "| gene-term pairs:", nrow(term2gene_bp), "\n")
}
```

### Coverage improvement

```{r}
#| label: eggnog-coverage

if (RUN_EGGNOG) {

  tested_genes <- unique(c(res_body_dev$gene_id, res_head_dev$gene_id,
                            res_body_flu$gene_id, res_head_flu$gene_id))

  n_tested      <- length(tested_genes)
  n_with_symbol <- sum(is_real_symbol(
                     gene_annot$gene_name[match(tested_genes, gene_annot$gene_id)]))
  n_with_go     <- dplyr::n_distinct(
                     dplyr::filter(term2gene_bp,
                                   gene_id %in% tested_genes)$gene_id)

  coverage_tbl <- tibble::tibble(
    Approach = c("Section 10 — symbol proxy (org.Hs.eg.db)",
                 "Section 10b — eggNOG GO terms (direct)"),
    Genes    = c(n_with_symbol, n_with_go),
    `% of tested` = round(100 * Genes / n_tested, 1)
  )

  knitr::kable(coverage_tbl,
               caption = "Enrichment-eligible genes: symbol proxy vs eggNOG direct annotation")
}
```

```{r}
#| label: eggnog-enrichment-fn

if (RUN_EGGNOG) {

  run_enrichment_eggnog <- function(res_df, label) {

    out <- list()

    for (dir in c("up", "down")) {

      universe_ids <- res_df |>
        dplyr::filter(!is.na(padj)) |>
        dplyr::pull(gene_id) |>
        intersect(unique(term2gene_bp$gene_id))

      sig_ids <- res_df |>
        dplyr::filter(
          !is.na(padj), padj < ALPHA,
          if (dir == "up") log2FoldChange >= LFC_CUTOFF
          else             log2FoldChange <= -LFC_CUTOFF) |>
        dplyr::pull(gene_id) |>
        intersect(unique(term2gene_bp$gene_id))

      if (length(sig_ids) < 5) {
        message(sprintf("%s / %s: %d genes with GO terms — skipped",
                        label, dir, length(sig_ids)))
        next
      }

      ego <- tryCatch(
        clusterProfiler::enricher(
          gene          = sig_ids,
          universe      = universe_ids,
          TERM2GENE     = term2gene_bp,
          TERM2NAME     = term2name_bp,
          pAdjustMethod = "BH",
          pvalueCutoff  = 0.05,
          qvalueCutoff  = 0.20
        ),
        error = function(e) {
          message(sprintf("enricher failed (%s/%s): %s", label, dir, conditionMessage(e)))
          NULL
        })

      if (is.null(ego) || nrow(as.data.frame(ego)) == 0) {
        message(sprintf("%s / %s: no enriched BP terms", label, dir))
        next
      }

      df <- dplyr::mutate(as.data.frame(ego), contrast = label, direction = dir)
      readr::write_csv(df,
        file.path(OUT_DIR, "enrichment",
                  paste0(label, "_", dir, "_eggnog_GOBP.csv")))

      out[[dir]] <- list(result = ego, df = df)
    }

    out
  }

  enrich_eggnog <- list(
    leg_emergence_body  = run_enrichment_eggnog(res_body_dev, "leg_emergence_body"),
    nervous_system_head = run_enrichment_eggnog(res_head_dev, "nervous_system_head"),
    AR_blockade_body    = run_enrichment_eggnog(res_body_flu, "AR_blockade_body"),
    AR_blockade_head    = run_enrichment_eggnog(res_head_flu, "AR_blockade_head"),
    old_tissue_adjusted = run_enrichment_eggnog(res_old_df,   "old_tissue_adjusted")
  )
  enrich_eggnog <- enrich_eggnog[lengths(enrich_eggnog) > 0]

  cat("Contrasts with at least one enriched direction:",
      length(enrich_eggnog), "\n")
}
```

### Results

```{r}
#| label: eggnog-dotplots
#| fig-height: 7
#| fig-width: 9

if (RUN_EGGNOG && length(enrich_eggnog) > 0) {

  for (contrast_name in names(enrich_eggnog)) {
    directions <- enrich_eggnog[[contrast_name]]
    for (dir in names(directions)) {
      ego <- directions[[dir]]$result
      p <- clusterProfiler::dotplot(ego, showCategory = 15) +
        ggplot2::labs(
          title    = gsub("_", " ", contrast_name),
          subtitle = paste("Direction:", dir,
                           "| Genes in universe:", length(ego@universe))) +
        ggplot2::theme_bw(base_size = 10) +
        ggplot2::theme(axis.text.y = ggplot2::element_text(size = 8))

      filename <- paste0(contrast_name, "_", dir, "_eggnog_dotplot.png")
      ggplot2::ggsave(file.path(OUT_DIR, "figures", filename),
                      plot = p, width = 9, height = 7, dpi = 300)
      print(p)
    }
  }
}
```

```{r}
#| label: eggnog-summary-table

if (RUN_EGGNOG) {

  eggnog_dfs <- lapply(enrich_eggnog, function(x)
    dplyr::bind_rows(lapply(x, `[[`, "df")))

  if (length(eggnog_dfs) > 0) {
    all_eggnog <- dplyr::bind_rows(eggnog_dfs) |>
      dplyr::select(contrast, direction, Description, GeneRatio, BgRatio,
                    p.adjust, Count) |>
      dplyr::arrange(contrast, direction, p.adjust)

    readr::write_csv(all_eggnog,
      file.path(OUT_DIR, "enrichment", "all_contrasts_eggnog_GOBP.csv"))

    ## top 5 terms per contrast × direction for the inline table
    top_terms <- all_eggnog |>
      dplyr::group_by(contrast, direction) |>
      dplyr::slice_min(p.adjust, n = 5, with_ties = FALSE) |>
      dplyr::ungroup()

    knitr::kable(top_terms, digits = 4,
                 caption = "Top 5 enriched GO:BP terms per contrast and direction (eggNOG annotation)")
  }
}
```


## 10c. Additional enrichment: KEGG pathways, GSEA, and COG categories

This section runs three further analyses on the same eggNOG annotation. All three depend on
the objects built in Section 10b (`RUN_EGGNOG`, `emapper`, `term2gene_bp`, `term2name_bp`),
so Section 10b must run first.

```{r}
#| label: extra-enrich-setup

RUN_EXTRA <- RUN_EGGNOG && exists("term2gene_bp") && nrow(term2gene_bp) > 0
if (!RUN_EXTRA) message("Section 10c requires Section 10b to have run successfully.")
```

### KEGG pathway over-representation

```{r}
#| label: kegg-build

if (RUN_EXTRA) {

  ## TERM2GENE from eggNOG KEGG_Pathway column (values like "map04010,map04151")
  term2gene_kegg <- emapper |>
    dplyr::filter(!is.na(KEGG_Pathway)) |>
    dplyr::mutate(pathway = strsplit(KEGG_Pathway, ",")) |>
    tidyr::unnest(pathway) |>
    dplyr::mutate(pathway = trimws(pathway)) |>
    dplyr::filter(grepl("^map", pathway)) |>
    dplyr::select(pathway, gene_id) |>
    dplyr::distinct()

  ## Fetch human-readable pathway names from KEGG REST API.
  ## Falls back to pathway IDs if the network is unreachable.
  term2name_kegg <- tryCatch({
    raw <- readLines("https://rest.kegg.jp/list/pathway", warn = FALSE)
    tibble::tibble(raw = raw) |>
      tidyr::separate(raw, into = c("pathway", "term_name"), sep = "\t", extra = "drop") |>
      dplyr::mutate(pathway = sub("^path:", "", pathway))
  }, error = function(e) {
    message("KEGG REST API unreachable — pathway IDs used as labels.")
    tibble::tibble(pathway = unique(term2gene_kegg$pathway),
                   term_name = unique(term2gene_kegg$pathway))
  })

  cat("KEGG pathways:", dplyr::n_distinct(term2gene_kegg$pathway),
      "| gene-pathway pairs:", nrow(term2gene_kegg), "\n")
}
```

```{r}
#| label: kegg-ora

if (RUN_EXTRA) {

  run_kegg_ora <- function(res_df, label) {
    out <- list()
    for (dir in c("up", "down")) {
      universe_ids <- res_df |>
        dplyr::filter(!is.na(padj)) |>
        dplyr::pull(gene_id) |>
        intersect(unique(term2gene_kegg$gene_id))

      sig_ids <- res_df |>
        dplyr::filter(!is.na(padj), padj < ALPHA,
                      if (dir == "up") log2FoldChange >= LFC_CUTOFF
                      else             log2FoldChange <= -LFC_CUTOFF) |>
        dplyr::pull(gene_id) |>
        intersect(unique(term2gene_kegg$gene_id))

      if (length(sig_ids) < 5) next

      ek <- tryCatch(
        clusterProfiler::enricher(
          gene          = sig_ids,
          universe      = universe_ids,
          TERM2GENE     = term2gene_kegg,
          TERM2NAME     = term2name_kegg,
          pAdjustMethod = "BH",
          pvalueCutoff  = 0.05,
          qvalueCutoff  = 0.20),
        error = function(e) NULL)

      if (is.null(ek) || nrow(as.data.frame(ek)) == 0) next

      df <- dplyr::mutate(as.data.frame(ek), contrast = label, direction = dir)
      readr::write_csv(df,
        file.path(OUT_DIR, "enrichment", paste0(label, "_", dir, "_KEGG.csv")))
      out[[dir]] <- list(result = ek, df = df)
    }
    out
  }

  kegg_ora <- list(
    leg_emergence_body  = run_kegg_ora(res_body_dev, "leg_emergence_body"),
    nervous_system_head = run_kegg_ora(res_head_dev, "nervous_system_head"),
    AR_blockade_body    = run_kegg_ora(res_body_flu, "AR_blockade_body"),
    AR_blockade_head    = run_kegg_ora(res_head_flu, "AR_blockade_head"),
    old_tissue_adjusted = run_kegg_ora(res_old_df,   "old_tissue_adjusted")
  )
  kegg_ora <- kegg_ora[lengths(kegg_ora) > 0]
  cat("Contrasts with KEGG results:", length(kegg_ora), "\n")
}
```

```{r}
#| label: kegg-plots
#| fig-height: 6
#| fig-width: 9

if (RUN_EXTRA && length(kegg_ora) > 0) {

  for (contrast_name in names(kegg_ora)) {
    for (dir in names(kegg_ora[[contrast_name]])) {
      ek <- kegg_ora[[contrast_name]][[dir]]$result
      p <- clusterProfiler::dotplot(ek, showCategory = 15) +
        ggplot2::labs(title    = gsub("_", " ", contrast_name),
                      subtitle = paste("KEGG pathway ORA —", dir)) +
        ggplot2::theme_bw(base_size = 10) +
        ggplot2::theme(axis.text.y = ggplot2::element_text(size = 8))
      ggplot2::ggsave(
        file.path(OUT_DIR, "figures",
                  paste0(contrast_name, "_", dir, "_KEGG_dotplot.png")),
        plot = p, width = 9, height = 6, dpi = 300)
      print(p)
    }
  }

  all_kegg <- dplyr::bind_rows(lapply(kegg_ora, function(x)
    dplyr::bind_rows(lapply(x, `[[`, "df"))))
  readr::write_csv(all_kegg,
    file.path(OUT_DIR, "enrichment", "all_contrasts_KEGG.csv"))
}
```

### GSEA: GO Biological Process and KEGG

GSEA ranks *all* tested genes by their shrunken log2 fold change and tests whether genes
annotated with each term cluster at the top or bottom of that list. It does not use a
significance cutoff, making it more sensitive to coordinated pathway shifts where individual
genes fall below the DE threshold.

```{r}
#| label: gsea-fn

if (RUN_EXTRA) {

  ## Build a sorted named vector of shrunken LFCs for GSEA.
  ## NA LFCs (outliers, zero-count genes) are dropped; all remaining tested genes
  ## are included so the background is the full experiment, not just DE genes.
  make_ranks <- function(res_df) {
    res_df |>
      dplyr::filter(!is.na(log2FoldChange)) |>
      dplyr::arrange(dplyr::desc(log2FoldChange)) |>
      dplyr::select(gene_id, log2FoldChange) |>
      dplyr::distinct(gene_id, .keep_all = TRUE) |>
      tibble::deframe()
  }

  run_gsea <- function(ranks, term2gene, term2name, label, suffix) {
    gs <- tryCatch(
      clusterProfiler::GSEA(
        geneList      = ranks,
        TERM2GENE     = term2gene,
        TERM2NAME     = term2name,
        minGSSize     = 10,
        maxGSSize     = 500,
        pvalueCutoff  = 0.05,
        pAdjustMethod = "BH",
        eps           = 0,
        seed          = TRUE,
        verbose       = FALSE),
      error = function(e) {
        message(sprintf("GSEA failed (%s/%s): %s", label, suffix, conditionMessage(e)))
        NULL
      })

    if (is.null(gs) || nrow(as.data.frame(gs)) == 0) {
      message(sprintf("%s / %s: no significant GSEA terms", label, suffix))
      return(NULL)
    }

    df <- dplyr::mutate(as.data.frame(gs), contrast = label, gene_set = suffix)
    readr::write_csv(df,
      file.path(OUT_DIR, "enrichment", paste0(label, "_GSEA_", suffix, ".csv")))
    gs
  }

  contrasts_list <- list(
    leg_emergence_body  = res_body_dev,
    nervous_system_head = res_head_dev,
    AR_blockade_body    = res_body_flu,
    AR_blockade_head    = res_head_flu,
    old_tissue_adjusted = res_old_df
  )

  gsea_go   <- lapply(names(contrasts_list), function(nm)
    run_gsea(make_ranks(contrasts_list[[nm]]), term2gene_bp,   term2name_bp,   nm, "GOBP"))
  gsea_kegg <- lapply(names(contrasts_list), function(nm)
    run_gsea(make_ranks(contrasts_list[[nm]]), term2gene_kegg, term2name_kegg, nm, "KEGG"))

  names(gsea_go)   <- names(contrasts_list)
  names(gsea_kegg) <- names(contrasts_list)

  gsea_go   <- gsea_go[!vapply(gsea_go,   is.null, logical(1))]
  gsea_kegg <- gsea_kegg[!vapply(gsea_kegg, is.null, logical(1))]

  cat("GSEA GO results:   ", length(gsea_go),   "contrasts with significant terms\n")
  cat("GSEA KEGG results: ", length(gsea_kegg), "contrasts with significant terms\n")
}
```

```{r}
#| label: gsea-plots
#| fig-height: 6
#| fig-width: 9

if (RUN_EXTRA) {

  plot_gsea <- function(gs_list, suffix) {
    for (nm in names(gs_list)) {
      gs <- gs_list[[nm]]
      p <- clusterProfiler::dotplot(gs, showCategory = 15, split = ".sign") +
        ggplot2::facet_grid(. ~ .sign) +
        ggplot2::labs(title    = gsub("_", " ", nm),
                      subtitle = paste("GSEA —", suffix)) +
        ggplot2::theme_bw(base_size = 10) +
        ggplot2::theme(axis.text.y = ggplot2::element_text(size = 8))
      ggplot2::ggsave(
        file.path(OUT_DIR, "figures", paste0(nm, "_GSEA_", suffix, "_dotplot.png")),
        plot = p, width = 9, height = 6, dpi = 300)
      print(p)
    }
  }

  if (length(gsea_go)   > 0) plot_gsea(gsea_go,   "GOBP")
  if (length(gsea_kegg) > 0) plot_gsea(gsea_kegg, "KEGG")
}
```

### COG functional category profile

COG categories describe the *broad functional class* of each gene model. This is not a
statistical enrichment test, it shows whether upregulated and downregulated genes are
concentrated in particular functional classes compared with all tested genes.

```{r}
#| label: cog-profile
#| fig-cap: "COG functional category distribution. Each bar shows the proportion of genes in that category among all tested genes (grey), upregulated DE genes (red), and downregulated DE genes (blue). Categories with fewer than 10 tested genes are omitted."
#| fig-height: 7
#| fig-width: 10

if (RUN_EXTRA) {

  COG_LABELS <- c(
    A = "RNA processing", B = "Chromatin structure",
    C = "Energy metabolism", D = "Cell cycle / division",
    E = "Amino acid metabolism", F = "Nucleotide metabolism",
    G = "Carbohydrate metabolism", H = "Coenzyme metabolism",
    I = "Lipid metabolism", J = "Translation / ribosome",
    K = "Transcription", L = "DNA replication / repair",
    M = "Cell wall / membrane", N = "Cell motility",
    O = "Protein turnover / chaperones", P = "Ion transport / metabolism",
    Q = "Secondary metabolites", R = "General function only",
    S = "Function unknown", T = "Signal transduction",
    U = "Vesicular transport", V = "Defense mechanisms",
    W = "Extracellular structures", Z = "Cytoskeleton"
  )

  ## one row per COG letter per gene (a gene can have multiple letters)
  cog_genes <- emapper |>
    dplyr::filter(!is.na(COG_category), COG_category != "-") |>
    dplyr::mutate(cog = strsplit(COG_category, "")) |>
    tidyr::unnest(cog) |>
    dplyr::filter(cog %in% names(COG_LABELS)) |>
    dplyr::select(gene_id, cog) |>
    dplyr::distinct()

  ## use AR_blockade_body as the representative contrast; swap for another if preferred
  focal <- res_body_flu |>
    dplyr::filter(!is.na(padj)) |>
    dplyr::mutate(de_class = dplyr::case_when(
      padj < ALPHA & log2FoldChange >=  LFC_CUTOFF ~ "up",
      padj < ALPHA & log2FoldChange <= -LFC_CUTOFF ~ "down",
      TRUE ~ "background"))

  cog_counts <- cog_genes |>
    dplyr::inner_join(focal |> dplyr::select(gene_id, de_class), by = "gene_id") |>
    dplyr::count(cog, de_class) |>
    dplyr::group_by(de_class) |>
    dplyr::mutate(prop = n / sum(n)) |>
    dplyr::ungroup() |>
    dplyr::mutate(
      label = paste0(cog, ": ", COG_LABELS[cog]),
      de_class = factor(de_class, levels = c("up", "background", "down")))

  ## keep only categories with enough tested-gene representation
  keep_cogs <- cog_counts |>
    dplyr::filter(de_class == "background", n >= 10) |>
    dplyr::pull(cog)

  p_cog <- cog_counts |>
    dplyr::filter(cog %in% keep_cogs) |>
    ggplot2::ggplot(ggplot2::aes(prop, reorder(label, prop), fill = de_class)) +
    ggplot2::geom_col(position = "dodge", width = 0.7) +
    ggplot2::scale_fill_manual(
      values = c(up = "#C44E52", background = "grey75", down = "#4C72B0"),
      labels = c(up = "Up-regulated", background = "All tested", down = "Down-regulated")) +
    ggplot2::scale_x_continuous(labels = scales::percent_format(accuracy = 1)) +
    ggplot2::labs(
      x = "Proportion of genes in category",
      y = NULL, fill = NULL,
      title = "COG functional categories — AR blockade body (old flutamide vs control)",
      caption = "Categories with < 10 tested genes omitted") +
    ggplot2::theme_bw(base_size = 10) +
    ggplot2::theme(legend.position = "top",
                   axis.text.y = ggplot2::element_text(size = 8))

  ggplot2::ggsave(file.path(OUT_DIR, "figures", "COG_profile_AR_blockade_body.png"),
                  plot = p_cog, width = 10, height = 7, dpi = 300)
  print(p_cog)

  readr::write_csv(
    cog_counts |> dplyr::select(cog, label, de_class, n, prop),
    file.path(OUT_DIR, "enrichment", "COG_profile_AR_blockade_body.csv"))
}
```

## 10d. Export gene ID lists for promoter motif enrichment (HPC)

This section exports the gene ID lists needed by `run_ame_motif.sbatch` on the cluster.
Run this locally after the DESeq2 contrasts are complete, then scp the output folder to the HPC.

```{r}
#| label: export-gene-lists

GENELIST_DIR <- file.path(OUT_DIR, "gene_lists_for_ame")
dir.create(GENELIST_DIR, recursive = TRUE, showWarnings = FALSE)

## all five contrasts as a named list
contrast_results <- list(
  leg_emergence_body  = res_body_dev,
  nervous_system_head = res_head_dev,
  AR_blockade_body    = res_body_flu,
  AR_blockade_head    = res_head_flu,
  old_tissue_adjusted = res_old_df
)

for (cname in names(contrast_results)) {

  df <- contrast_results[[cname]]

  ## universe: all genes that received a padj (i.e. not filtered by independent filtering)
  universe_ids <- df |>
    dplyr::filter(!is.na(padj)) |>
    dplyr::pull(gene_id)

  ## up-regulated: padj < ALPHA, log2FC >= LFC_CUTOFF
  up_ids <- df |>
    dplyr::filter(!is.na(padj), padj < ALPHA, log2FoldChange >= LFC_CUTOFF) |>
    dplyr::pull(gene_id)

  ## down-regulated: padj < ALPHA, log2FC <= -LFC_CUTOFF
  down_ids <- df |>
    dplyr::filter(!is.na(padj), padj < ALPHA, log2FoldChange <= -LFC_CUTOFF) |>
    dplyr::pull(gene_id)

  readr::write_lines(universe_ids, file.path(GENELIST_DIR, paste0(cname, "_universe_geneids.txt")))
  readr::write_lines(up_ids,       file.path(GENELIST_DIR, paste0(cname, "_up_geneids.txt")))
  readr::write_lines(down_ids,     file.path(GENELIST_DIR, paste0(cname, "_down_geneids.txt")))

  cat(sprintf("%-30s  universe=%d  up=%d  down=%d\n",
              cname, length(universe_ids), length(up_ids), length(down_ids)))
}

cat("\nGenerate lists written to:", GENELIST_DIR, "\n")
cat("\nUpload to HPC with:\n")
cat(sprintf("  scp -r %s kfloer_smith_edu@login.smith.edu:/scratch3/workspace/kfloer_smith_edu-simple/motif_analysis/gene_lists\n",
            GENELIST_DIR))
```

## 11. Reproducibility

```{r}
#| label: session

writeLines(capture.output(sessionInfo()),
           file.path(OUT_DIR, "sessionInfo.txt"))

write_csv(coldata, file.path(OUT_DIR, "tables", "sample_metadata.csv"))

cat("Outputs written to:", normalizePath(OUT_DIR), "\n\n")
print(list.files(OUT_DIR, recursive = TRUE))

sessionInfo()
```