---
title: "Differential Gene Expression Analysis"
subtitle: "DESeq2 analysis with design, QC, and sensitivity audits"
format:
  html:
    code-fold: true
    code-tools: true
editor: visual
execute:
  eval: false
  warning: true
  message: true
---

## Scope and inferential limits

This analysis contains head and body RNA-seq libraries from one frog clutch. Day-45 tadpoles were collected before the tank treatment period. At day 103, tadpoles were sampled from one control tank and one flutamide tank.

Important limits are built into the analysis and must accompany interpretation:

- Treatment is confounded with tank because there is one tank per condition. The treatment-assignment unit is the tank, so tadpoles within a tank are subsamples rather than independent treatment replicates. Results are therefore described as **flutamide-tank versus control-tank associations**, not definitive causal effects of flutamide.
- The experiment is not a complete tissue × treatment × stage factorial: no young flutamide samples exist. Overall treatment effects across stages, treatment-by-stage interactions, and three-way interactions are not estimable.
- All tadpoles came from one clutch, so inference is clutch-specific.
- Stage comparisons use different animals collected on different days. They combine age/development with any day- or cohort-specific differences.
- Only two old-control head libraries exist. Head contrasts and the tissue-by-treatment interaction are consequently fragile.

## Setup

```{r}
#| label: setup
library(DESeq2)
library(tximport)
library(ggplot2)
library(ggrepel)
library(pheatmap)
library(matrixStats)

if (!requireNamespace("ashr", quietly = TRUE)) {
  stop("Install 'ashr' with BiocManager::install('ashr').")
}
if (!requireNamespace("rtracklayer", quietly = TRUE)) {
  stop("Install 'rtracklayer' with BiocManager::install('rtracklayer').")
}

dir.create("DESeq2_results", showWarnings = FALSE)
dir.create("DESeq2_results/QC", showWarnings = FALSE)
dir.create("DESeq2_results/DGE", showWarnings = FALSE)
dir.create("DESeq2_results/volcano", showWarnings = FALSE)
dir.create("DESeq2_results/sensitivity", showWarnings = FALSE)
```

## Authoritative sample manifest

The manifest is explicit rather than inferred solely from filenames. `tank` is recorded for provenance but cannot be added beside treatment in a model because the two are perfectly confounded at day 103.

```{r}
#| label: sample-manifest
sample_manifest <- data.frame(
  sample_id = c(
    "C103-2B_S10_L002", "C103-3B_S11_L002", "C103-4B_S13_L002",
    "C103-4H_S12_L002", "C103-5B_S15_L002", "C103-5H_S14_L002",
    "C45-1B_S2_L002", "C45-1H_S1_L002", "C45-2B_S4_L002",
    "C45-2H_S3_L002", "C45-3B_S5_L002", "C45-4B_S7_L002",
    "C45-4H_S6_L002", "C45-5B_S9_L002", "C45-5H_S8_L002",
    "F103-1B_S17_L002", "F103-1H_S16_L002", "F103-2B_S19_L002",
    "F103-2H_S18_L002", "F103-3B_S21_L002", "F103-3H_S20_L002",
    "F103-4B_S22_L002", "F103-5B_S24_L002", "F103-5H_S23_L002"
  ),
  tadpole = c(
    "C103-2", "C103-3", "C103-4", "C103-4", "C103-5", "C103-5",
    "C45-1", "C45-1", "C45-2", "C45-2", "C45-3", "C45-4",
    "C45-4", "C45-5", "C45-5",
    "F103-1", "F103-1", "F103-2", "F103-2", "F103-3", "F103-3",
    "F103-4", "F103-5", "F103-5"
  ),
  day = c(rep(103L, 6), rep(45L, 9), rep(103L, 9)),
  treatment = c(rep("Control", 15), rep("Flutamide", 9)),
  tissue = c(
    "Body", "Body", "Body", "Head", "Body", "Head",
    "Body", "Head", "Body", "Head", "Body", "Body", "Head", "Body", "Head",
    "Body", "Head", "Body", "Head", "Body", "Head", "Body", "Body", "Head"
  ),
  clutch = "clutch_1",
  tank = c(rep("control_tank_1", 6), rep("pre_treatment", 9), rep("flutamide_tank_1", 9)),
  stringsAsFactors = FALSE
)

stopifnot(nrow(sample_manifest) == 24L)
stopifnot(!anyDuplicated(sample_manifest$sample_id))
stopifnot(!anyNA(sample_manifest))

sample_manifest$stage <- factor(
  ifelse(sample_manifest$day == 45L, "Young", "Old"),
  levels = c("Young", "Old")
)
sample_manifest$treatment <- factor(
  sample_manifest$treatment,
  levels = c("Control", "Flutamide")
)
sample_manifest$tissue <- factor(
  sample_manifest$tissue,
  levels = c("Body", "Head")
)
sample_manifest$group <- factor(
  paste(sample_manifest$stage, sample_manifest$treatment,
        sample_manifest$tissue, sep = "_"),
  levels = c(
    "Young_Control_Body", "Young_Control_Head",
    "Old_Control_Body", "Old_Control_Head",
    "Old_Flutamide_Body", "Old_Flutamide_Head"
  )
)

expected_group_counts <- c(
  Young_Control_Body = 5L, Young_Control_Head = 4L,
  Old_Control_Body = 4L, Old_Control_Head = 2L,
  Old_Flutamide_Body = 5L, Old_Flutamide_Head = 4L
)
observed_group_counts <- table(sample_manifest$group)
stopifnot(identical(as.integer(observed_group_counts), unname(expected_group_counts)))

sample_manifest
observed_group_counts
```

## Salmon import with tximport

DESeq2 recommends importing Salmon transcript quantifications with `tximport`, then constructing the dataset with `DESeqDataSetFromTximport()`. This retains gene-specific average transcript-length offsets. TPM values are not used for testing.

```{r}
#| label: tximport
quant_dir <- "salmon_quant"
tx2gene_file <- "salmon.merged.tx2gene.tsv"
gtf_file <- "complete.genomic.gtf"

files <- file.path(quant_dir, sample_manifest$sample_id, "quant.sf")
names(files) <- sample_manifest$sample_id
stopifnot(all(file.exists(files)))
stopifnot(length(files) == 24L)

tx2gene_raw <- read.delim(
  tx2gene_file,
  header = TRUE,
  stringsAsFactors = FALSE,
  check.names = FALSE
)
stopifnot(all(c("transcript_id", "gene_id") %in% names(tx2gene_raw)))
tx2gene <- unique(tx2gene_raw[, c("transcript_id", "gene_id")])
stopifnot(!anyNA(tx2gene), !anyDuplicated(tx2gene$transcript_id))

txi <- tximport(
  files,
  type = "salmon",
  tx2gene = tx2gene,
  countsFromAbundance = "no"
)
stopifnot(identical(colnames(txi$counts), sample_manifest$sample_id))

input_checksums <- data.frame(
  file = c(files, tx2gene_file, gtf_file),
  md5 = unname(tools::md5sum(c(files, tx2gene_file, gtf_file))),
  stringsAsFactors = FALSE
)
write.csv(
  input_checksums,
  "DESeq2_results/input_file_md5_checksums.csv",
  row.names = FALSE
)
```

## Gene annotation

```{r}
#| label: gene-annotation
gene_annotation <- unique(tx2gene_raw[, c("gene_id", "gene")])
names(gene_annotation)[2] <- "gene_name"

gtf <- rtracklayer::import(gtf_file)
gtf_mcols <- mcols(gtf)
stopifnot(all(c("gene_id", "type", "product") %in% names(gtf_mcols)))

gtf_products <- data.frame(
  gene_id = as.character(gtf_mcols$gene_id),
  feature_type = as.character(gtf_mcols$type),
  product = as.character(gtf_mcols$product),
  stringsAsFactors = FALSE
)
gtf_products <- subset(
  gtf_products,
  !is.na(gene_id) & gene_id != "" & !is.na(product) & product != ""
)
product_annotation <- do.call(rbind, lapply(
  split(gtf_products, gtf_products$gene_id),
  function(x) {
    preferred <- unique(x$product[x$feature_type %in% c("transcript", "mRNA")])
    products <- if (length(preferred)) preferred else unique(x$product)
    data.frame(
      gene_id = x$gene_id[1],
      product_description = paste(products, collapse = " | "),
      stringsAsFactors = FALSE
    )
  }
))
rownames(product_annotation) <- NULL
rm(gtf, gtf_mcols, gtf_products)
```

## DESeq2 construction and fit

The observed cell combinations are represented by one `group` factor. This is the DESeq2 manual's recommended simple strategy for explicit cell contrasts and avoids a non-estimable full factorial model.

```{r}
#| label: construct-and-fit
coldata <- sample_manifest
rownames(coldata) <- coldata$sample_id

dds <- DESeqDataSetFromTximport(
  txi,
  colData = coldata,
  design = ~ group
)
stopifnot(identical(rownames(colData(dds)), colnames(dds)))
stopifnot(qr(model.matrix(design(dds), as.data.frame(colData(dds))))$rank ==
            nlevels(dds$group))

mcols(dds)$gene_name <- gene_annotation$gene_name[
  match(rownames(dds), gene_annotation$gene_id)
]
mcols(dds)$product_description <- product_annotation$product_description[
  match(rownames(dds), product_annotation$gene_id)
]

# Intentional information filter: count >=10 in at least 3 libraries.
# Three is stricter than the actual smallest cell (old-control head, n=2).
genes_before_filter <- nrow(dds)
keep <- rowSums(counts(dds) >= 10) >= 3
dds <- dds[keep, ]
filter_audit <- data.frame(
  genes_before_filter = genes_before_filter,
  genes_after_filter = nrow(dds),
  genes_removed = genes_before_filter - nrow(dds),
  rule = "count >= 10 in at least 3 libraries"
)
write.csv(filter_audit, "DESeq2_results/prefilter_audit.csv", row.names = FALSE)
filter_audit

dds <- DESeq(dds)
resultsNames(dds)

write.csv(as.data.frame(colData(dds)),
          "DESeq2_results/sample_metadata.csv", row.names = FALSE)
saveRDS(dds, "DESeq2_results/dds_fitted.rds")
```

## Quality control

### Library sizes, effective scaling factors, and zero fractions

```{r}
#| label: library-qc
library_qc <- data.frame(
  sample_id = colnames(dds),
  sample_core = sub("_S[0-9]+_L[0-9]+$", "", colnames(dds)),
  group = dds$group,
  raw_gene_count_sum = colSums(counts(dds)),
  # tximport supplies gene-by-sample normalization factors that combine
  # average transcript-length offsets with library scaling. Their column
  # geometric means provide a sample-level scaling summary for QC only.
  effective_library_scaling = exp(
    colMeans(log(normalizationFactors(dds)))
  ),
  zero_fraction = colMeans(counts(dds) == 0),
  stringsAsFactors = FALSE
)
write.csv(library_qc, "DESeq2_results/QC/library_size_factor_zero_fraction.csv",
          row.names = FALSE)

library_qc_plot <- ggplot(
  library_qc,
  aes(x = reorder(sample_core, effective_library_scaling),
      y = effective_library_scaling, fill = group)
) +
  geom_col() + coord_flip() +
  labs(title = "tximport/DESeq2 effective library scaling", x = NULL,
       y = "Geometric mean of normalization factors", fill = "Group") +
  theme_classic(base_size = 11)
ggsave("DESeq2_results/QC/effective_library_scaling.pdf", library_qc_plot,
       width = 10, height = 7)
library_qc_plot
```

Small effective library-scaling values are flags for upstream MultiQC review,
not automatic exclusion criteria. Because tximport supplies gene-specific
length offsets, these summaries must not be described as ordinary DESeq2 size
factors.

### Dispersion and Cook's-distance diagnostics

```{r}
#| label: dispersion-cooks-qc
pdf("DESeq2_results/QC/dispersion_estimates.pdf", width = 8, height = 6)
plotDispEsts(dds)
dev.off()

cooks <- assays(dds)[["cooks"]]
cooks_summary <- data.frame(
  sample_id = colnames(dds),
  median_cooks = apply(cooks, 2, median, na.rm = TRUE),
  q99_cooks = apply(cooks, 2, quantile, probs = 0.99, na.rm = TRUE),
  max_cooks = apply(cooks, 2, max, na.rm = TRUE)
)
write.csv(cooks_summary, "DESeq2_results/QC/cooks_distance_by_sample.csv",
          row.names = FALSE)

pdf("DESeq2_results/QC/cooks_distance_boxplot.pdf", width = 11, height = 7)
par(mar = c(9, 5, 2, 2))
boxplot(log10(cooks), range = 0, las = 2,
        ylab = "log10 Cook's distance", main = "Cook's distance by library")
dev.off()
```

### VST, PCA, sample distances, and heatmaps

`blind=FALSE` uses the fitted dispersion trend for biological visualization and does not remove group effects. A second `blind=TRUE` PCA is included as design-blind sample QA.

```{r}
#| label: transformation-pca
vsd <- vst(dds, blind = FALSE)
vsd_blind <- vst(dds, blind = TRUE)
saveRDS(vsd, "DESeq2_results/vsd.rds")

make_pca <- function(vsd_object, label, filename) {
  mat <- assay(vsd_object)
  vars <- rowVars(mat)
  top <- names(sort(setNames(vars, rownames(mat)), decreasing = TRUE))[
    seq_len(min(500, nrow(mat)))
  ]
  pca <- prcomp(t(mat[top, , drop = FALSE]), scale. = FALSE)
  percent <- 100 * pca$sdev^2 / sum(pca$sdev^2)
  pd <- cbind(as.data.frame(pca$x[, 1:2, drop = FALSE]),
              as.data.frame(colData(vsd_object)))
  p <- ggplot(pd, aes(PC1, PC2, color = group, shape = tissue,
                     label = sub("_S[0-9]+_L[0-9]+$", "", sample_id))) +
    geom_point(size = 3.5) +
    geom_text_repel(size = 2.8, max.overlaps = Inf, show.legend = FALSE) +
    labs(title = paste("PCA:", label),
         subtitle = "500 most variable genes",
         x = sprintf("PC1 (%.1f%%)", percent[1]),
         y = sprintf("PC2 (%.1f%%)", percent[2]), color = "Group") +
    coord_fixed() + theme_classic(base_size = 11)
  ggsave(filename, p, width = 10, height = 7)
  list(plot = p, data = pd, top = top, variance = vars)
}

pca_bio <- make_pca(vsd, "VST, blind=FALSE",
                    "DESeq2_results/QC/PCA_top500_variable_genes.pdf")
pca_blind <- make_pca(vsd_blind, "VST, blind=TRUE",
                      "DESeq2_results/QC/PCA_top500_variable_genes_blind.pdf")
write.csv(pca_bio$data, "DESeq2_results/QC/PCA_sample_coordinates.csv",
          row.names = FALSE)
write.csv(data.frame(
  gene_id = pca_bio$top,
  gene_name = mcols(dds)[pca_bio$top, "gene_name"],
  VST_variance = pca_bio$variance[pca_bio$top]
), "DESeq2_results/QC/top500_variable_genes.csv", row.names = FALSE)
pca_bio$plot
pca_blind$plot
```

```{r}
#| label: distance-heatmaps
vst_matrix <- assay(vsd)
top500 <- pca_bio$top
sample_dist <- dist(t(vst_matrix[top500, , drop = FALSE]))
sample_dist_matrix <- as.matrix(sample_dist)
annotation_col <- as.data.frame(colData(vsd)[, c("stage", "treatment", "tissue", "group")])

pheatmap(sample_dist_matrix, annotation_col = annotation_col,
         annotation_row = annotation_col,
         clustering_distance_rows = sample_dist,
         clustering_distance_cols = sample_dist,
         filename = "DESeq2_results/QC/sample_distance_heatmap.pdf",
         width = 11, height = 10,
         main = "Sample distances: top 500 variable genes")

top50 <- top500[seq_len(min(50, length(top500)))]
heatmap_z <- t(scale(t(vst_matrix[top50, , drop = FALSE])))
heatmap_z[!is.finite(heatmap_z)] <- 0
labels <- mcols(dds)[top50, "gene_name"]
labels <- ifelse(is.na(labels) | labels == "", top50,
                 paste0(labels, " [", top50, "]"))
rownames(heatmap_z) <- make.unique(labels)
pheatmap(heatmap_z, annotation_col = annotation_col,
         fontsize_row = 6, fontsize_col = 7,
         color = colorRampPalette(c("navy", "white", "firebrick3"))(101),
         breaks = seq(-3, 3, length.out = 102),
         filename = "DESeq2_results/QC/top50_variable_gene_heatmap.pdf",
         width = 12, height = 12,
         main = "Top 50 variable genes (row z-scores)")
```

## Differential-expression functions

The standard Wald result tests LFC = 0. A separate thresholded Wald result formally tests whether |LFC| > 1. Shrunken LFCs are used for ranking and visualization, not to create p-values.

```{r}
#| label: comparison-functions
annotate_result <- function(res, dds_object) {
  out <- as.data.frame(res)
  out$gene_id <- rownames(out)
  out$gene_name <- mcols(dds_object)[out$gene_id, "gene_name"]
  out$product_description <- mcols(dds_object)[out$gene_id, "product_description"]
  out
}

run_comparison <- function(dds, numerator, denominator, comparison_name,
                           alpha = 0.05, lfc_threshold = 1) {
  contrast_vector <- c("group", numerator, denominator)
  res <- results(dds, contrast = contrast_vector, alpha = alpha)
  res_threshold <- results(
    dds, contrast = contrast_vector, alpha = alpha,
    lfcThreshold = lfc_threshold, altHypothesis = "greaterAbs"
  )
  res_shrunk <- lfcShrink(
    dds, contrast = contrast_vector, res = res, type = "ashr"
  )

  out <- annotate_result(res, dds)
  out$shrunken_log2FoldChange <- res_shrunk$log2FoldChange
  out$shrunken_lfcSE <- res_shrunk$lfcSE
  out$threshold_stat <- res_threshold$stat
  out$threshold_pvalue <- res_threshold$pvalue
  out$threshold_padj <- res_threshold$padj
  out$standard_FDR_significant <- !is.na(out$padj) & out$padj < alpha
  out$LFC_threshold_FDR_significant <-
    !is.na(out$threshold_padj) & out$threshold_padj < alpha
  out$threshold_direction <- "Not threshold-significant"
  out$threshold_direction[
    out$LFC_threshold_FDR_significant & out$shrunken_log2FoldChange > 0
  ] <- "Higher"
  out$threshold_direction[
    out$LFC_threshold_FDR_significant & out$shrunken_log2FoldChange < 0
  ] <- "Lower"
  out <- out[order(out$threshold_padj, out$padj, out$pvalue, na.last = TRUE), ]

  write.csv(out, file.path("DESeq2_results/DGE",
    paste0(comparison_name, "_DESeq2_results.csv")), row.names = FALSE)

  result_summary <- data.frame(
    comparison = comparison_name,
    numerator = numerator,
    denominator = denominator,
    genes = nrow(out),
    standard_FDR_below_0.05 = sum(out$standard_FDR_significant),
    threshold_test_FDR_below_0.05 = sum(out$LFC_threshold_FDR_significant),
    threshold_higher = sum(out$threshold_direction == "Higher"),
    threshold_lower = sum(out$threshold_direction == "Lower"),
    NA_pvalue = sum(is.na(out$pvalue)),
    NA_padj = sum(is.na(out$padj)),
    independent_filter_threshold = unname(metadata(res)$filterThreshold)
  )

  pdf(file.path("DESeq2_results/QC", paste0(comparison_name, "_MA.pdf")),
      width = 8, height = 6)
  plotMA(res_shrunk, alpha = alpha, ylim = c(-5, 5),
         main = paste(comparison_name, "(ashr LFC)"))
  dev.off()

  plot_data <- out
  plot_data$plot_padj <- ifelse(is.na(plot_data$threshold_padj), 1,
                               pmax(plot_data$threshold_padj,
                                    .Machine$double.xmin))
  labels <- subset(plot_data,
                   LFC_threshold_FDR_significant & !is.na(gene_name) & gene_name != "")
  labels <- head(labels[order(labels$threshold_padj), ], 10)
  volcano <- ggplot(plot_data,
    aes(shrunken_log2FoldChange, -log10(plot_padj), color = threshold_direction)) +
    geom_point(alpha = 0.65, size = 1.4) +
    geom_vline(xintercept = c(-lfc_threshold, lfc_threshold),
               linetype = "dashed", color = "grey40") +
    geom_hline(yintercept = -log10(alpha), linetype = "dashed", color = "grey40") +
    geom_text_repel(data = labels, aes(label = gene_name), size = 3,
                    max.overlaps = Inf, show.legend = FALSE) +
    scale_color_manual(values = c(
      "Lower" = "#2C7BB6", "Not threshold-significant" = "grey75",
      "Higher" = "#D7191C"
    )) +
    labs(title = comparison_name,
         subtitle = paste(numerator, "vs", denominator),
         x = "ashr-shrunken log2 fold change",
         y = expression(-log[10]("threshold-test adjusted p-value")),
         caption = paste0("Formal Wald test of |LFC| > ", lfc_threshold,
                          "; FDR < ", alpha), color = NULL) +
    theme_classic(base_size = 11) + theme(legend.position = "top")
  ggsave(file.path("DESeq2_results/volcano",
                   paste0(comparison_name, "_volcano.pdf")),
         volcano, width = 8, height = 7)

  list(res = res, res_threshold = res_threshold, res_shrunk = res_shrunk,
       table = out, summary = result_summary)
}
```

## Primary contrasts

```{r}
#| label: primary-contrasts
leg_emergence <- run_comparison(
  dds, "Old_Control_Body", "Young_Control_Body",
  "leg_emergence_OldControlBody_vs_YoungControlBody"
)
nervous_system_development <- run_comparison(
  dds, "Old_Control_Head", "Young_Control_Head",
  "nervous_system_OldControlHead_vs_YoungControlHead"
)
ar_blockade_body <- run_comparison(
  dds, "Old_Flutamide_Body", "Old_Control_Body",
  "AR_blockade_OldBodyFlutamide_vs_OldBodyControl"
)
ar_blockade_head <- run_comparison(
  dds, "Old_Flutamide_Head", "Old_Control_Head",
  "AR_blockade_OldHeadFlutamide_vs_OldHeadControl"
)

contrast_summaries <- do.call(rbind, lapply(list(
  leg_emergence, nervous_system_development,
  ar_blockade_body, ar_blockade_head
), `[[`, "summary"))
write.csv(contrast_summaries, "DESeq2_results/DGE/contrast_summaries.csv",
          row.names = FALSE)
contrast_summaries
```

The stage contrasts are control-only, between-animal day/stage comparisons. The flutamide contrasts are day-103 flutamide-tank versus control-tank associations. The old-head contrasts have only two control libraries and are considered provisional.

## Relevant-subset sensitivity fits

DESeq2 usually recommends fitting all groups together, but very different groups can influence the shared gene-wise dispersion. Each main contrast is therefore refit using only its two relevant cells.

```{r}
#| label: subset-sensitivity
run_subset_sensitivity <- function(dds_full, full_result, numerator,
                                   denominator, comparison_name) {
  use <- dds_full$group %in% c(numerator, denominator)
  d <- dds_full[, use]
  d$group <- droplevels(d$group)
  design(d) <- ~ group
  d <- DESeq(d)
  sub_res <- results(d, contrast = c("group", numerator, denominator), alpha = 0.05)
  ids <- rownames(full_result)
  sub_lfc <- sub_res$log2FoldChange[match(ids, rownames(sub_res))]
  sub_padj <- sub_res$padj[match(ids, rownames(sub_res))]
  tab <- data.frame(
    gene_id = ids,
    full_log2FC = full_result$log2FoldChange,
    full_padj = full_result$padj,
    subset_log2FC = sub_lfc,
    subset_padj = sub_padj,
    direction_agrees = sign(full_result$log2FoldChange) == sign(sub_lfc)
  )
  finite <- is.finite(tab$full_log2FC) & is.finite(tab$subset_log2FC)
  summary <- data.frame(
    comparison = comparison_name,
    full_samples = ncol(dds_full), subset_samples = ncol(d),
    spearman_LFC = cor(tab$full_log2FC[finite], tab$subset_log2FC[finite],
                       method = "spearman"),
    direction_agreement = mean(tab$direction_agrees[finite], na.rm = TRUE),
    full_FDR_hits = sum(tab$full_padj < 0.05, na.rm = TRUE),
    subset_FDR_hits = sum(tab$subset_padj < 0.05, na.rm = TRUE),
    shared_FDR_hits = sum(tab$full_padj < 0.05 & tab$subset_padj < 0.05,
                          na.rm = TRUE)
  )
  write.csv(tab, file.path("DESeq2_results/sensitivity",
    paste0(comparison_name, "_full_vs_subset.csv")), row.names = FALSE)
  summary
}

subset_summaries <- rbind(
  run_subset_sensitivity(dds, leg_emergence$res,
    "Old_Control_Body", "Young_Control_Body", "stage_body"),
  run_subset_sensitivity(dds, nervous_system_development$res,
    "Old_Control_Head", "Young_Control_Head", "stage_head"),
  run_subset_sensitivity(dds, ar_blockade_body$res,
    "Old_Flutamide_Body", "Old_Control_Body", "flutamide_body"),
  run_subset_sensitivity(dds, ar_blockade_head$res,
    "Old_Flutamide_Head", "Old_Control_Head", "flutamide_head")
)
write.csv(subset_summaries,
          "DESeq2_results/sensitivity/full_vs_subset_summaries.csv",
          row.names = FALSE)
subset_summaries
```

## Older-only paired tissue interaction

The interaction is

$$
(\mathrm{flutamide\ tank\ head}-\mathrm{control\ tank\ head})-
(\mathrm{flutamide\ tank\ body}-\mathrm{control\ tank\ body}).
$$

Tadpole fixed effects account for paired head/body libraries. Treatment main effect is absorbed by tadpole; the custom `flutamide_head` coefficient estimates the difference-in-differences.

```{r}
#| label: paired-interaction
dds_old <- dds[, dds$stage == "Old"]
# Use syntactically safe factor levels for stable coefficient names.
dds_old$tadpole <- factor(make.names(as.character(dds_old$tadpole)))
dds_old$tissue <- factor(dds_old$tissue, levels = c("Body", "Head"))
dds_old$flutamide_head <- as.integer(
  dds_old$treatment == "Flutamide" & dds_old$tissue == "Head"
)
design(dds_old) <- ~ tadpole + tissue + flutamide_head
paired_mm <- model.matrix(design(dds_old), as.data.frame(colData(dds_old)))
stopifnot(qr(paired_mm)$rank == ncol(paired_mm))

pair_table <- with(as.data.frame(colData(dds_old)), table(tadpole, tissue))
complete_pair_ids <- rownames(pair_table)[rowSums(pair_table > 0) == 2]
paired_design_audit <- data.frame(
  old_libraries = ncol(dds_old),
  tadpoles = nlevels(dds_old$tadpole),
  complete_pairs = length(complete_pair_ids),
  control_complete_pairs = sum(grepl("^C", complete_pair_ids)),
  flutamide_complete_pairs = sum(grepl("^F", complete_pair_ids)),
  model_parameters = ncol(paired_mm),
  residual_df = nrow(paired_mm) - qr(paired_mm)$rank
)
write.csv(paired_design_audit,
          "DESeq2_results/DGE/paired_interaction_design_audit.csv",
          row.names = FALSE)
paired_design_audit

dds_old <- DESeq(dds_old)
interaction_name <- grep("flutamide_head", resultsNames(dds_old), value = TRUE)
stopifnot(length(interaction_name) == 1L)
res_interaction <- results(dds_old, name = interaction_name, alpha = 0.05)
res_interaction_shrunk <- lfcShrink(
  dds_old, coef = interaction_name, res = res_interaction, type = "ashr"
)

interaction_out <- annotate_result(res_interaction, dds_old)
interaction_out$shrunken_interaction_log2FC <-
  res_interaction_shrunk$log2FoldChange
interaction_out$shrunken_interaction_lfcSE <- res_interaction_shrunk$lfcSE

# Auxiliary component contrasts come from the separate global, unpaired model.
interaction_out$auxiliary_unpaired_head_log2FC <-
  ar_blockade_head$res$log2FoldChange[
    match(interaction_out$gene_id, rownames(ar_blockade_head$res))
  ]
interaction_out$auxiliary_unpaired_body_log2FC <-
  ar_blockade_body$res$log2FoldChange[
    match(interaction_out$gene_id, rownames(ar_blockade_body$res))
  ]
interaction_out$paired_interaction_FDR_significant <-
  !is.na(interaction_out$padj) & interaction_out$padj < 0.05
interaction_out <- interaction_out[
  order(interaction_out$padj, interaction_out$pvalue, na.last = TRUE),
]
write.csv(interaction_out,
          "DESeq2_results/DGE/AR_blockade_head_vs_body_interaction_old_only.csv",
          row.names = FALSE)
saveRDS(dds_old, "DESeq2_results/dds_old_interaction_fitted.rds")
```

## Unpaired interaction sensitivity analysis

```{r}
#| label: unpaired-interaction
dds_old_unpaired <- dds[, dds$stage == "Old"]
dds_old_unpaired$treatment <- factor(dds_old_unpaired$treatment,
                                     levels = c("Control", "Flutamide"))
dds_old_unpaired$tissue <- factor(dds_old_unpaired$tissue,
                                  levels = c("Body", "Head"))
design(dds_old_unpaired) <- ~ treatment * tissue
unpaired_mm <- model.matrix(design(dds_old_unpaired),
                            as.data.frame(colData(dds_old_unpaired)))
stopifnot(qr(unpaired_mm)$rank == ncol(unpaired_mm))
dds_old_unpaired <- DESeq(dds_old_unpaired)
unpaired_name <- grep("treatment.*tissue|tissue.*treatment",
                      resultsNames(dds_old_unpaired), value = TRUE)
stopifnot(length(unpaired_name) == 1L)
res_unpaired <- results(dds_old_unpaired, name = unpaired_name, alpha = 0.05)
res_unpaired_shrunk <- lfcShrink(
  dds_old_unpaired, coef = unpaired_name, res = res_unpaired, type = "ashr"
)

sensitivity_comparison <- data.frame(
  gene_id = rownames(res_interaction),
  gene_name = mcols(dds_old)[rownames(res_interaction), "gene_name"],
  paired_interaction_log2FC = res_interaction_shrunk$log2FoldChange,
  paired_interaction_padj = res_interaction$padj,
  unpaired_interaction_log2FC = res_unpaired_shrunk$log2FoldChange[
    match(rownames(res_interaction), rownames(res_unpaired_shrunk))
  ],
  unpaired_interaction_padj = res_unpaired$padj[
    match(rownames(res_interaction), rownames(res_unpaired))
  ]
)
sensitivity_comparison$direction_agrees <- with(
  sensitivity_comparison,
  sign(paired_interaction_log2FC) == sign(unpaired_interaction_log2FC)
)
sensitivity_comparison$paired_significant <- with(
  sensitivity_comparison, !is.na(paired_interaction_padj) &
    paired_interaction_padj < 0.05
)
sensitivity_comparison$unpaired_significant <- with(
  sensitivity_comparison, !is.na(unpaired_interaction_padj) &
    unpaired_interaction_padj < 0.05
)
finite <- with(sensitivity_comparison,
  is.finite(paired_interaction_log2FC) & is.finite(unpaired_interaction_log2FC))
interaction_sensitivity_summary <- data.frame(
  genes_compared = sum(finite),
  spearman_LFC = with(sensitivity_comparison,
    cor(paired_interaction_log2FC[finite], unpaired_interaction_log2FC[finite],
        method = "spearman")),
  direction_agreement = mean(sensitivity_comparison$direction_agrees[finite],
                             na.rm = TRUE),
  paired_FDR_hits = sum(sensitivity_comparison$paired_significant),
  unpaired_FDR_hits = sum(sensitivity_comparison$unpaired_significant),
  shared_FDR_hits = with(sensitivity_comparison,
    sum(paired_significant & unpaired_significant))
)
write.csv(sensitivity_comparison,
  "DESeq2_results/DGE/AR_blockade_interaction_paired_vs_unpaired_all_genes.csv",
  row.names = FALSE)
write.csv(interaction_sensitivity_summary,
  "DESeq2_results/DGE/AR_blockade_interaction_paired_vs_unpaired_summary.csv",
  row.names = FALSE)
saveRDS(dds_old_unpaired,
        "DESeq2_results/dds_old_unpaired_interaction_fitted.rds")
interaction_sensitivity_summary
```

## Leave-one-complete-pair-out stability

Each complete head/body pair is omitted in turn. This does not create new experimental replication; it tests whether leading paired-interaction estimates are dominated by one pair.

```{r}
#| label: leave-one-pair-out
paired_hit_ids <- rownames(res_interaction)[
  !is.na(res_interaction$padj) & res_interaction$padj < 0.05
]
full_hit_lfc <- res_interaction$log2FoldChange[
  match(paired_hit_ids, rownames(res_interaction))
]
loo_list <- lapply(complete_pair_ids, function(id) {
  d <- dds_old[, dds_old$tadpole != id]
  d$tadpole <- droplevels(d$tadpole)
  design(d) <- ~ tadpole + tissue + flutamide_head
  mm <- model.matrix(design(d), as.data.frame(colData(d)))
  stopifnot(qr(mm)$rank == ncol(mm))
  d <- DESeq(d, quiet = TRUE)
  nm <- grep("flutamide_head", resultsNames(d), value = TRUE)
  r <- results(d, name = nm, alpha = 0.05)
  data.frame(
    omitted_pair = id,
    gene_id = paired_hit_ids,
    full_log2FC = full_hit_lfc,
    leave_one_out_log2FC = r$log2FoldChange[
      match(paired_hit_ids, rownames(r))
    ],
    leave_one_out_padj = r$padj[match(paired_hit_ids, rownames(r))]
  )
})
loo_stability <- do.call(rbind, loo_list)
loo_stability$direction_agrees <- with(
  loo_stability, sign(full_log2FC) == sign(leave_one_out_log2FC)
)
write.csv(loo_stability,
          "DESeq2_results/sensitivity/paired_interaction_leave_one_pair_out.csv",
          row.names = FALSE)
loo_summary <- aggregate(
  cbind(direction_agrees, leave_one_out_log2FC) ~ gene_id,
  data = loo_stability,
  FUN = function(x) c(min = min(x, na.rm = TRUE),
                      median = median(x, na.rm = TRUE),
                      max = max(x, na.rm = TRUE))
)
loo_summary
```

## Targeted spinal cord motor-neuron and interneuron marker audit

This section checks the candidate list supplied by the PI. Antibody/common
names are resolved to the gene symbols used in the *S. parvus* annotation.
Grouped markers are expanded to individual paralogs. "Gone4" is interpreted
as **GON4L**; Hb9 as **MNX1**; Chx10 as **VSX2**; HuC/HuD as
**ELAVL3/ELAVL4**; and Raldh2 as **ALDH1A2**.

```{r}
#| label: candidate-gene-map
candidate_map <- data.frame(
  PI_marker = c(
    "Hb9", "Lhx3 (Lim3)", "Isl1/2", "Isl1/2", "foxp1", "En1",
    "Evx1", "dmrt3", "nkx2.2", "Chx10", "Shox2", "Hu C/D", "Hu C/D",
    "Sox2", "Gone4", "Lmx1-2", "Lmx1-2", "MafA", "Otp", "foxp2",
    "Sox1/3", "Sox1/3", "Sp8", "Lhx2-9 (BamhII)",
    "Lhx2-9 (BamhII)", "Raldh2", "ChAT"
  ),
  resolved_symbol = c(
    "MNX1", "LHX3", "ISL1", "ISL2", "FOXP1", "EN1", "EVX1", "DMRT3",
    "NKX2-2", "VSX2", "SHOX2", "ELAVL3", "ELAVL4", "SOX2", "GON4L",
    "LMX1A", "LMX1B", "MAFA", "OTP", "FOXP2", "SOX1", "SOX3", "SP8",
    "LHX2", "LHX9", "ALDH1A2", "CHAT"
  ),
  alias_note = c(
    "Hb9 is MNX1", "Lim3 is LHX3", "Isl1", "Isl2", "", "", "", "",
    "NKX2-2", "Chx10 is VSX2", "", "HuC is ELAVL3", "HuD is ELAVL4",
    "", "Gone4 interpreted as GON4L", "Lmx1a", "Lmx1b", "", "", "",
    "Sox1", "Sox3", "", "Lhx2", "Lhx9", "Raldh2 is ALDH1A2", ""
  ),
  stringsAsFactors = FALSE
)

annotation_lookup <- unique(tx2gene_raw[, c("gene_id", "gene")])
candidate_map$gene_id <- annotation_lookup$gene_id[
  match(candidate_map$resolved_symbol, annotation_lookup$gene)
]
candidate_map$present_in_annotation <- !is.na(candidate_map$gene_id)
candidate_map$retained_after_filter <-
  candidate_map$gene_id %in% rownames(dds)

stopifnot(all(candidate_map$present_in_annotation))
candidate_map
```

Presence in the annotation does not guarantee sufficient expression for
differential testing. A gene absent after filtering had fewer than three
libraries with counts of at least 10 and is reported as "not tested," not as
evidence of no expression.

```{r}
#| label: candidate-gene-results
candidate_comparisons <- list(
  stage_body = leg_emergence,
  stage_head = nervous_system_development,
  flutamide_tank_body = ar_blockade_body,
  flutamide_tank_head = ar_blockade_head
)

candidate_long <- do.call(rbind, lapply(names(candidate_comparisons), function(nm) {
  obj <- candidate_comparisons[[nm]]
  tab <- obj$table
  idx <- match(candidate_map$gene_id, tab$gene_id)
  data.frame(
    PI_marker = candidate_map$PI_marker,
    resolved_symbol = candidate_map$resolved_symbol,
    gene_id = candidate_map$gene_id,
    comparison = nm,
    present_in_annotation = candidate_map$present_in_annotation,
    retained_after_filter = candidate_map$retained_after_filter,
    baseMean = tab$baseMean[idx],
    shrunken_log2FoldChange = tab$shrunken_log2FoldChange[idx],
    pvalue = tab$pvalue[idx],
    padj = tab$padj[idx],
    threshold_padj = tab$threshold_padj[idx],
    standard_FDR_significant = tab$standard_FDR_significant[idx],
    LFC_threshold_FDR_significant =
      tab$LFC_threshold_FDR_significant[idx],
    stringsAsFactors = FALSE
  )
}))

# Add the paired tissue interaction as a fifth comparison. It tests whether
# the day-103 between-tank association differs between head and body.
interaction_idx <- match(candidate_map$gene_id, interaction_out$gene_id)
candidate_interaction <- data.frame(
  PI_marker = candidate_map$PI_marker,
  resolved_symbol = candidate_map$resolved_symbol,
  gene_id = candidate_map$gene_id,
  comparison = "paired_tissue_interaction",
  present_in_annotation = candidate_map$present_in_annotation,
  retained_after_filter = candidate_map$retained_after_filter,
  baseMean = interaction_out$baseMean[interaction_idx],
  shrunken_log2FoldChange =
    interaction_out$shrunken_interaction_log2FC[interaction_idx],
  pvalue = interaction_out$pvalue[interaction_idx],
  padj = interaction_out$padj[interaction_idx],
  threshold_padj = NA_real_,
  standard_FDR_significant =
    interaction_out$paired_interaction_FDR_significant[interaction_idx],
  LFC_threshold_FDR_significant = NA,
  stringsAsFactors = FALSE
)
candidate_long <- rbind(candidate_long, candidate_interaction)

candidate_long$direction <- ifelse(
  is.na(candidate_long$shrunken_log2FoldChange), NA,
  ifelse(candidate_long$shrunken_log2FoldChange > 0, "Higher", "Lower")
)
candidate_long$DE_call <- ifelse(
  !candidate_long$retained_after_filter, "Not tested: low counts",
  ifelse(is.na(candidate_long$padj), "Tested; adjusted p-value unavailable",
    ifelse(candidate_long$standard_FDR_significant,
           paste("FDR < 0.05", candidate_long$direction),
           "Not significant at FDR 0.05"))
)

write.csv(
  candidate_long,
  "DESeq2_results/DGE/spinal_marker_candidates_all_comparisons.csv",
  row.names = FALSE
)

candidate_overview <- do.call(rbind, lapply(seq_len(nrow(candidate_map)), function(i) {
  z <- candidate_long[candidate_long$resolved_symbol ==
                        candidate_map$resolved_symbol[i], ]
  data.frame(
    PI_marker = candidate_map$PI_marker[i],
    resolved_symbol = candidate_map$resolved_symbol[i],
    gene_id = candidate_map$gene_id[i],
    present_in_annotation = candidate_map$present_in_annotation[i],
    retained_after_filter = candidate_map$retained_after_filter[i],
    any_standard_FDR_hit = any(z$standard_FDR_significant, na.rm = TRUE),
    significant_comparisons = paste(
      z$comparison[which(z$standard_FDR_significant %in% TRUE)],
      collapse = "; "
    ),
    any_formal_abs_LFC_above_1_hit =
      any(z$LFC_threshold_FDR_significant, na.rm = TRUE),
    stringsAsFactors = FALSE
  )
}))
write.csv(
  candidate_overview,
  "DESeq2_results/DGE/spinal_marker_candidates_presence_summary.csv",
  row.names = FALSE
)
candidate_overview
```

The standard FDR column answers whether a candidate differs from zero under
any comparison. The formal effect-threshold column is more stringent and asks
whether an absolute LFC greater than 1 is statistically supported. The paired
interaction is evaluated only against zero, because the interaction analysis
was not specified as an effect-threshold test.

```{r}
#| label: candidate-gene-expression
candidate_ids <- candidate_map$gene_id[candidate_map$retained_after_filter]
candidate_symbols <- candidate_map$resolved_symbol[
  candidate_map$retained_after_filter
]
normalized_candidate_counts <- counts(dds, normalized = TRUE)[
  candidate_ids, , drop = FALSE
]

candidate_expression_long <- do.call(rbind, lapply(seq_along(candidate_ids), function(i) {
  data.frame(
    gene_id = candidate_ids[i],
    resolved_symbol = candidate_symbols[i],
    sample_id = colnames(dds),
    group = dds$group,
    normalized_count = normalized_candidate_counts[i, ],
    stringsAsFactors = FALSE
  )
}))
write.csv(
  candidate_expression_long,
  "DESeq2_results/DGE/spinal_marker_candidates_normalized_counts.csv",
  row.names = FALSE
)

candidate_expression_plot <- ggplot(
  candidate_expression_long,
  aes(x = group, y = log2(normalized_count + 1), color = group)
) +
  geom_boxplot(outlier.shape = NA, alpha = 0.15, color = "grey45") +
  geom_point(position = position_jitter(width = 0.12), size = 0.9,
             alpha = 0.75) +
  facet_wrap(~ resolved_symbol, scales = "free_y", ncol = 4) +
  labs(
    title = "Targeted spinal cord marker expression across bulk tissue groups",
    subtitle = "DESeq2-normalized gene counts; points are libraries",
    x = NULL,
    y = "log2(normalized count + 1)",
    color = "Group"
  ) +
  theme_classic(base_size = 9) +
  theme(
    axis.text.x = element_text(angle = 60, hjust = 1, size = 6),
    legend.position = "none"
  )
ggsave(
  "DESeq2_results/QC/spinal_marker_candidates_normalized_counts.pdf",
  candidate_expression_plot,
  width = 15,
  height = 17
)
candidate_expression_plot
```

These measurements come from bulk head and body dissections. A marker can be
strongly localized by immunohistochemistry yet show no bulk-tissue DE if its
cell population is rare, if opposing cell populations dilute the signal, or
if expression changes spatially without changing the tissue-wide mean.

## Interpretation

A positive interaction means the flutamide-tank association is more positive in head than body; it may arise from stronger head induction, stronger body repression, or both. A negative interaction has the opposite meaning. Interaction sign is not ordinary gene up- or downregulation.

The most defensible interaction candidates are genes with consistent effect directions, stability when individual complete pairs are omitted, and support in both paired and unpaired models. Statistical interaction does not establish direct androgen-receptor regulation and remains confounded with tank.

Suggested reporting language:

> At day 103, tadpoles housed in the flutamide tank differed in gene expression from tadpoles housed in the control tank. Because treatment was applied at the tank level with one tank per condition, treatment and tank effects are inseparable and these contrasts are exploratory. The tissue-dependent association was estimated from six complete head–body pairs (two control and four flutamide; paired-model residual df = 4). Candidate genes were evaluated using paired and unpaired interaction models and leave-one-pair-out stability analyses.

## Reproducibility record

```{r}
#| label: reproducibility
writeLines(capture.output(sessionInfo()), "DESeq2_results/sessionInfo.txt")
sessionInfo()
```
